---
title: "Testing some models for seal data"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

```{r}
#| warning: false
#| message: false

library(ggplot2); theme_set(theme_light())
library(stringr)
library(tidyr)
library(dplyr)
library(tidylog)
library(ggeffects)
library(sdmTMB)
library(modelr)
library(readr)

home <- here::here()
```

Read data

```{r}
#| warning: false
#| message: false

d <- read_csv(paste0(home, "/data/HS_SKAG_matrix_biomass_abundance_filtered_ok.csv")) |> 
  janitor::clean_names() |> 
  rename(pw = sum_biomass_not_corrected) |> 
  mutate(family = as.factor(family),
         year_f = as.factor(year),
         year_ct = year - min(year),
         month_f = as.factor(month)) |> 
  filter(lat > 57)
```

Now expand the data so that every predator has one row for each prey type (currently no zeroes-by-prey group)

```{r}
#| warning: false
#| message: false

# Expand the data: If a seal stomach contained Gadidae but not Clupeidae, we add a row for Clupeidae with pw = 0. This is necessary because we want to model the probability of eating each prey type AND how much of it.
d <- d |>
  group_by(sample_id) |>
  complete(
    family,
    fill = list(
      pw = 0, 
      sum_abundance = 0
    )
  ) |>
  fill(lat, lon, year, month, year_f, year_ct, month_f,
       .direction = "downup") |> 
  ungroup()

# Check if we need to summarise prey weight a bit more
d |> 
  summarise(sum_pw = sum(pw), .by = family) |> 
  mutate(percent = (sum_pw / sum(sum_pw)) * 100) |> 
  arrange(desc(percent))

# Group all prey below 3% into "other"
d <- d |>
  mutate(percent = 100 * sum(pw) / sum(d$pw), .by = family) |>
  mutate(family = if_else(percent < 3,
                          "Other",
                          as.character(family))) |>
  select(-percent) |>
  summarise(
    pw = sum(pw),
    sum_abundance = sum(sum_abundance),
    .by = c(sample_id, family, lat, lon, year,
            month, year_f, year_ct, month_f)
  )
```

Explore the data quickly

```{r}
#| warning: false
#| message: false

head(d)

ggplot(d, aes(lon, lat, color = as.factor(year))) + 
  geom_point() + 
  coord_sf()

d |> 
  summarise(n = n_distinct(sample_id), .by = year) |> 
  ggplot(aes(year, n)) + 
  geom_bar(stat = "identity")

summary(d$pw)

# FIXME
d <- d |> 
  filter(pw < 10000)
# > d |> 
# +   filter(pw > 14000)
# filter: removed 1,294 rows (>99%), one row remaining# A tibble: 1 × 11
#   sample_id   family      lat   lon  year month year_f year_ct month_f     pw sum_abundance
#   <chr>       <chr>     <dbl> <dbl> <dbl> <dbl> <fct>    <dbl> <fct>    <dbl>         <dbl>
# 1 A2020/05589 Clupeidae  58.9  11.1  2016     1 2016         5 1       19748.            35

ggplot(d, aes(pw)) + 
  geom_histogram()

ggplot(d, aes(month)) + 
  geom_histogram()
```

All models use the R package sdmTMB (Species Distribution Modelling with TMB). Think of it as a powerful toolbox that combines:

GLMs (Generalized Linear Models) — for modelling non-normal data
GAMs (Generalized Additive Models) — for smooth, nonlinear effects
Spatial random fields — for capturing spatial patterns (like kriging in geostatistics)
Time series — AR1 autoregressive processes

Prey weight is a tricky variable: it has many zeros (seal didn't eat that prey family) and the positive values are right-skewed (most are small, a few are very large). Two approaches are used:

Delta-gamma (delta_gamma(type = "poisson-link")): Fits two sub-models simultaneously:

A binomial model for the probability of presence (did the seal eat this prey? yes/no)
A gamma model for the positive amount (how much, given it ate it)
The "poisson-link" means both sub-models share a log link, so their predictions can be simply added on the link scale.

Tweedie (tweedie(link = "log")): A single distribution that naturally handles zeros and positive continuous values. Simpler but less flexible than delta-gamma.

Fit a simple model, estimating the mean prey weight by family, ignoring space and time


```{r}
#| warning: false
#| message: false

#m1: Baseline — "What is the average diet?"
# Fit a simple model
# Let's use a delta_gamma model to have a bit more flexibility than a tweedie
m1 <- sdmTMB(
  pw ~ 0 + family, # prey weight predicted by family only
  data = d,
  family = delta_gamma(type = "poisson-link"),
  spatial = "off"  # no spatial effects
)

summary(m1)

nd <- d |> 
  distinct(family)

# est here is the combined prediction from the binomial and the gamma model
# It's the same as est1 + est2. This is one of the reasons we use a type = "poisson-link",
# because in standard delta models the two sub models have different link functions and thereby
# cannot be added
p <- predict(m1, newdata = nd, se_fit = TRUE) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

ggplot(p, aes(exp(est), reorder(family, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Check residuals of the combined model (both components). We do this with dharma
res <- simulate(m1, nsim = 300, type = "mle-mvn") |>
  dharma_residuals(m1, plot = FALSE)

# Plot residuals
res |> 
  ggplot(aes(observed, expected)) +
  geom_point(color = "grey30", shape = 21, size = 0.5) +
  geom_abline(col = "tomato3", linewidth = 0.6) +
  theme(aspect.ratio = 1) +
  labs(x = "Observed", y = "Expected")

# It looks slightly better if you remove the uncommon families
```
Gadidae has the highest average prey weight, followed by Clupeidae and then "other". The confidence intervals are quite wide for some families, especially the less common ones.

m2: Adding year as a factor — "Does diet change between years?"
Now add year to the model. There are many options for how. We'll look into a factor year, linear year and autoregressive year effects.
It estimates a different overall prey level for each year, while keeping family effects. Each year gets its own intercept — no assumption about trend shape.
```{r}
#| warning: false
#| message: false

m2 <- sdmTMB(
  pw ~ 0 + family + year_f, # add a separate intercept for each year
  data = d,
  family = delta_gamma(type = "poisson-link"),
  spatial = "off"
)

nd2 <- d |> 
  distinct(family, year_f)

p2 <- predict(m2, newdata = nd2, se_fit = TRUE) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se,
         year = as.numeric(as.character(year_f)))

p2 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family) |> 
  ggplot(aes(exp(est), reorder(family, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction. The year-effect plot shows whether total prey 
# weight goes up or down in particular years
p2 |> 
  ggplot(aes(year, exp(est), color = family, fill = family)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family)

# And now the total prey weight over time. The family × year prediction plot 
# shows year-by-year trends for each prey family, while the total prey weight 
# plot shows the overall trend in prey weight over time, regardless of family.
p2 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = year) |> 
  ggplot(aes(year, exp(est))) + 
  geom_point() + 
  geom_errorbar(aes(ymin = exp(lwr), ymax = exp(upr)),
                alpha = 0.3, width = 0)
```

Linear year effect
m3: Linear year trend — "Is there a long-term trend?". Fits a single straight 
line through time. One slope parameter instead of one per year.
```{r}
#| warning: false
#| message: false

m3 <- sdmTMB(
  pw ~ family + year_ct, # use centered year to aid convergence
  data = d,
  family = delta_gamma(type = "poisson-link"),
  spatial = "off"
)

nd3 <- d |> 
  distinct(family, year_ct, year)

p3 <- predict(m3, newdata = nd3, se_fit = TRUE) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p3 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family) |> 
  ggplot(aes(exp(est), reorder(family, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p3 |> 
  ggplot(aes(year, exp(est), color = family, fill = family)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family) 

# And now the total prey weight over time
p3 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = year) |> 
  ggplot(aes(year, exp(est))) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA)
```

Linear year effect but different slopes for each family
m4: Family × year interaction — "Do different prey trend differently?" 
This is a more complex model that allows each prey family to have its own linear
trend over time. It includes an interaction between family and year, so we can 
see if some prey types are increasing while others are decreasing.
Allows each prey family to have a different linear trend over time. 
E.g., Gadidae could be increasing while Clupeidae is decreasing.
```{r}
#| warning: false
#| message: false

m4 <- sdmTMB(
  pw ~ family*year_ct, # use centered year to aid convergence
  data = d,
  family = delta_gamma(type = "poisson-link"),
  spatial = "off"
)

nd4 <- d |> 
  distinct(family, year_ct, year)

p4 <- predict(m4, newdata = nd4, se_fit = TRUE) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p4 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family) |> 
  ggplot(aes(exp(est), reorder(family, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p4 |> 
  ggplot(aes(year, exp(est), color = family, fill = family)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family) 

# And now the total prey weight over time
p4 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = year) |> 
  ggplot(aes(year, exp(est))) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA)
```

The interaction above explodes for some families..an interaction with few data 
points leads to huge uncertainty.

What if we wanted to model time for as a time series, using an autoregressive approach?
m5: AR1 time-varying intercept — "Are there autocorrelated year effects?"
Instead of treating each year independently (m2) or assuming a straight line (m3), 
it models the year effect as an autoregressive process: each year's effect is a 
function of the previous year's effect, plus some noise.
```{r}
#| warning: false
#| message: false

m5 <- sdmTMB(
  pw ~ 1 + family,
  data = d,
  time_varying = ~1,
  # TODO: we could also do a random walk, but ar1 seems easier to converge. basically: do we want the trend to be able to drift off
  # (only year before matters), or do we want the annual values to revert to mean in absence of data?
  time_varying_type = "ar1",
  # now that we are modelling a process over time, we need to specify any missing years we want to predict over, using process error only
  extra_time = 2023,
  # we also need to define which is our time-column, now that we have a temporal model
  time = "year",
  family = delta_gamma(type = "poisson-link"),
  # when we add a time-component, the default is to add a spatiotemporal field. We turn this off for now! 
  spatial = "off",
  spatiotemporal = "off"
)

summary(m5)

sanity(m5)

nd5 <- d |> 
  distinct(family, year) |> 
  # add the missing year, 2023!
  bind_rows(d |> 
              distinct(family) |> 
              mutate(year = 2023))
  
p5 <- predict(m5, newdata = nd5, se_fit = TRUE) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p5 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family) |> 
  ggplot(aes(exp(est), reorder(family, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p5 |> 
  ggplot(aes(year, exp(est), color = family, fill = family)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family) 

# And now the total prey weight over time
# Note how we get an estimate for 2023 now!
p5 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = year) |> 
  ggplot(aes(year, exp(est))) + 
  geom_point() + 
  geom_errorbar(aes(ymin = exp(lwr), ymax = exp(upr)),
                alpha = 0.3, width = NA)
```

```{r}
#| warning: false
#| message: false

AIC(m1, m2, m3, m4, m5) |> 
  arrange(AIC)
```

Now testing adding spatial random effects
m6: Adding spatial random effects — "Does diet vary by location?"
Adds a spatial random field (a "map" of random effects) to account for the fact
that seals in nearby locations tend to have similar diets. 
Uses a mesh (a triangulation of sampling locations) to define spatial structure.
```{r}
#| warning: false
#| message: false

d <- d |> 
  add_utm_columns(ll_names = c("lon", "lat"))

#The mesh is a triangulated network over the study area. 
#The model estimates a smooth spatial surface by estimating values at mesh 
#vertices and interpolating between them. cutoff = 4 means vertices must be at 
#least 4 km apart.

mesh <- make_mesh(d, xy_cols = c("X", "Y"), type = "cutoff", cutoff = 4)

plot(mesh)

mesh$mesh$n

m6 <- sdmTMB(
  pw ~ 0 + family + year_f,
  data = d,
  mesh = mesh,
  family = delta_gamma(type = "poisson-link"),
  # TODO: this can be explored a bit. From experience and from testing with this data a bit,
  # it's much easier to estimate the binomial part (including any random effects), compared to the positive biomass part
  # Hence I estimate here only a spatial random effect for the binomial part.
  spatial = list("on", "off"),
  # priors = sdmTMBpriors(
  #   matern_s = pc_matern(range_gt = 5, sigma_lt = 1)
  #   )
)

summary(m6)
sanity(m6)

nd6 <- d |> 
  distinct(family, year, year_f)

p6 <- predict(m6, newdata = nd6, se_fit = TRUE, re_form = NA) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p6 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family) |> 
  ggplot(aes(exp(est), reorder(family, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p6 |> 
  ggplot(aes(year, exp(est), color = family, fill = family)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family) 

# And now the total prey weight over time
p6 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = year) |> 
  ggplot(aes(year, exp(est))) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA)
```

Spatial random _and_ time-varying intercept
m7: Spatial + AR1 time-varying — "Spatial and temporal effects together"
Combines the spatial field from m6 with the AR1 temporal process from m5. 
The spatial field is constant across years — it captures permanent spatial patterns.

```{r}
#| warning: false
#| message: false

m7 <- sdmTMB(
  pw ~ 1 + family,
  data = d,
  time_varying = ~1,
  time_varying_type = "ar1",
  mesh = mesh,
  extra_time = 2023,
  time = "year",
  family = delta_gamma(type = "poisson-link"),
  spatial = list("on", "off"),
  spatiotemporal = "off"
)

# Plot spatial random effect?
nd7 <- d |>
  distinct(X, Y, lon, lat) |> 
  mutate(year = 2011, 
         year_f = as.factor("2011"),
         family = as.factor("Gadidae"))

p7 <- predict(m7, newdata = nd7, re_form_iid = NA)

p7

p7 |> 
  filter(lat > 57.7) |> 
  ggplot(aes(lon, lat, color = omega_s1)) + 
  geom_point(size = 2) + 
  scale_color_gradient2()
```

It's a pretty clear pattern here, that the amount of prey in the stomachs declines with distance from coast!

```{r}
# FIXME: this is the delta_gamma models!
AIC(m1, m2, m3, m4, m5, m6, m7) |> 
  arrange(AIC)
```


Now, we want the predation to change spatially with family. Spatial-varying effect of family!
m8: Spatially-varying family effects — "Do different areas favour different prey?"
Instead of one spatial field for all prey combined, each prey family gets its own spatial field. 
This models the idea that, say, Gadidae might be more important near the coast 
while Clupeidae dominates offshore.
```{r}
d$family_f <- as.factor(d$family)

m8 <- sdmTMB(
  pw ~ 0 + family_f + year_f,
  data = d,
  mesh = mesh,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(link = "log"),
  spatial = "off",
  spatial_varying = ~0 + family_f
)

nd8 <- d |>
  distinct(X, Y, lon, lat) |> 
  mutate(year_f = as.factor("2011")) |> 
  replicate_df(time_values = unique(d$family_f), time_name = "family_f")

p8 <- predict(m8, newdata = nd8)

names(p8)

p8 |> 
  pivot_longer(starts_with("zeta"), names_to = "var", values_to = "zeta") |> 
  ggplot(aes(lon, lat, color = zeta)) + 
  geom_point(size = 1) + 
  facet_wrap(~var) +
  scale_color_gradient2()

p8 |> 
  pivot_longer(starts_with("zeta"), names_to = "var", values_to = "zeta") |> 
  filter(var %in% c("zeta_s_family_fGadidae", "zeta_s_family_fClupeidae")) |> 
  ggplot(aes(lon, lat, color = zeta)) + 
  geom_point(size = 1) + 
  facet_wrap(~var) +
  scale_color_gradient2()

#The zeta columns are the spatially-varying coefficients. 
#Each zeta_s_family_fXXX is a map showing where that prey family's contribution 
#is positively or negatively shifted relative to the mean.
#Positive zeta = that family has higher-than-average prey weight in that area
#Negative zeta = lower-than-average
```

What if we do both time varying intercept and spatial varying family?
m9: Spatially-varying family + AR1 intercept — "Spatial and temporal effects together, with family-specific spatial patterns"
```{r}
m9 <- sdmTMB(
  pw ~ 1 + family_f,
  data = d,
  time_varying = ~1,
  time_varying_type = "ar1",
  mesh = mesh,
  extra_time = 2023,
  time = "year",
  family = tweedie(link = "log"),
  spatial = "off",
  spatial_varying = ~0 + family_f,
  spatiotemporal = "off"
)

m9
sanity(m9)

nd9 <- d |>
  distinct(X, Y, lon, lat) |> 
  mutate(year = 2011, 
         year_f = as.factor("2011")) |> 
  replicate_df(time_values = unique(d$family_f), time_name = "family_f") |> 
  droplevels()
  
p9 <- predict(m9, newdata = nd9, re_form_iid = NA)

names(p9)

p9 |> 
  pivot_longer(starts_with("zeta"), names_to = "var", values_to = "zeta") |> 
  ggplot(aes(lon, lat, color = zeta)) + 
  geom_point(size = 1) + 
  facet_wrap(~var) +
  scale_color_gradient2()

# Now family and time trends
nd9b <- d |> 
  distinct(family_f, year) |> 
  bind_rows(d |> 
              distinct(family_f) |> 
              mutate(year = 2023))
  
p9b <- predict(m9, newdata = nd9b, se_fit = TRUE, re_form = NA) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p9b |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family_f) |> 
  ggplot(aes(exp(est), reorder(family_f, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p9b |> 
  ggplot(aes(year, exp(est), color = family_f, fill = family_f)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family_f) 
```

One last thing to check: can we fit a model where the families have specific temporal trends?
m10: GAM smooth year trends by family + spatial-varying coefficients
```{r}
# simplest case is to have a year_f * family_f interaction... but its not efficient and doesnt converge!
d$year_sc <- (d$year - mean(d$year)) / sd(d$year)

m10 <- sdmTMB(
  pw ~ 0 + family_f + s(year_sc, by = family_f, k = 3),
  data = d,
  mesh = mesh,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(link = "log"),
  spatial = "off",
  spatial_varying = ~0 + family_f
)

sanity(m10)

nd10 <- d |>
  distinct(X, Y, lon, lat) |> 
  mutate(year = 2011, 
         year_sc = 0,
         year_f = as.factor("2011")) |> 
  replicate_df(time_values = unique(d$family_f), time_name = "family_f") |> 
  droplevels()
  
p10 <- predict(m10, newdata = nd10, re_form_iid = NA)

names(p10)

p10 |> 
  pivot_longer(starts_with("zeta"), names_to = "var", values_to = "zeta") |> 
  ggplot(aes(lon, lat, color = zeta)) + 
  geom_point(size = 1) + 
  facet_wrap(~var) +
  scale_color_gradient2()

# Now family and time trends
nd10b <- d |> 
  distinct(family_f, year) |> 
  bind_rows(d |> 
              distinct(family_f) |> 
              mutate(year = 2023)) |> 
  mutate(year_sc = (year - mean(d$year)) / sd(d$year))
  
p10b <- predict(m10, newdata = nd10b, se_fit = TRUE, re_form = NA) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p10b |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family_f) |> 
  ggplot(aes(exp(est), reorder(family_f, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p10b |> 
  ggplot(aes(year, exp(est), color = family_f, fill = family_f)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family_f)

p10b |> 
  ggplot(aes(year, exp(est), color = family_f, fill = family_f)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family_f, scales = "free")

p10b |> 
  ggplot(aes(year, exp(est), color = family_f, fill = family_f)) + 
  geom_line() + 
  facet_wrap(~family_f, scales = "free") 
```

m11: Family-specific AR1 time-varying effects + spatial
Each prey family gets its own time-varying intercept (AR1). So Gadidae can have a different temporal trajectory than Clupeidae.
```{r}
m11 <- sdmTMB(
  pw ~ 1 + family_f,
  data = d,
  time_varying = ~1 + family_f,
  time_varying_type = "ar1",
  mesh = mesh,
  extra_time = 2023,
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  #spatial_varying = ~0 + family_f,
  spatiotemporal = "off"
)

# Now family and time trends
nd11 <- d |> 
  distinct(family_f, year) |> 
  bind_rows(d |> 
              distinct(family_f) |> 
              mutate(year = 2023)) |> 
  mutate(year_sc = (year - mean(d$year)) / sd(d$year))
  
p11 <- predict(m11, newdata = nd11, se_fit = TRUE, re_form = NA) |> 
  mutate(upr = est + 1.96*est_se,
         lwr = est - 1.96*est_se)

p11 |> 
  summarise(est = mean(est),
            upr = mean(upr),
            lwr = mean(lwr), .by = family_f) |> 
  ggplot(aes(exp(est), reorder(family_f, exp(est)))) + 
  geom_point() + 
  geom_errorbar(aes(xmin = exp(lwr), xmax = exp(upr)),
                width = 0, alpha = 0.5)

# Now plot the year-prediction
p11 |> 
  ggplot(aes(year, exp(est), color = family_f, fill = family_f)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family_f)

p11 |> 
  ggplot(aes(year, exp(est), color = family_f, fill = family_f)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)),
              alpha = 0.3, color = NA) +
  facet_wrap(~family_f, scales = "free")
```

How do these spatially-varying models stand against m2, fitted as a tweedie?

```{r}
m2_tw <- sdmTMB(
  pw ~ 0 + family + year_f,
  data = d,
  family = tweedie(),
  spatial = "off"
)

AIC(m2_tw, m8, m9, m10, m11) |> 
  arrange(AIC)
```

## To do now

1. Make a regular grid, and make predictions from m8 or m9 (i think i prefer that one actually, even though it has +3 AIC...)

Question: What does the spatial map of predicted prey weight actually look like?

All previous spatial models made predictions at observed sampling locations only (the dots where seals were caught). This script creates predictions on a regular grid, producing smooth, map-like visualisations.

```{r}
#Create a regular prediction grid
#    We build a grid in UTM space covering the sampled extent, then
#    convert back to lon/lat for plotting.
# Define grid resolution (km in UTM)
grid_res <- 1  # 1 km resolution

# Grid in UTM
x_range <- range(d$X)
y_range <- range(d$Y)

grid_utm <- expand.grid(
  X = seq(x_range[1] - 2, x_range[2] + 2, by = grid_res),
  Y = seq(y_range[1] - 2, y_range[2] + 2, by = grid_res)
)

cat("Grid cells:", nrow(grid_utm), "\n")
# Convert UTM back to lon/lat for plotting
# sdmTMB uses zone from add_utm_columns; we reverse it
# The UTM zone is stored as an attribute
utm_zone <- attr(d, "crs")
if (is.null(utm_zone)) {
  # Fallback: determine from median longitude
  utm_zone <- floor((median(d$lon) + 180) / 6) + 1
  cat("Estimated UTM zone:", utm_zone, "\n")
}

# Use sf for back-conversion
if (requireNamespace("sf", quietly = TRUE)) {
  grid_sf <- sf::st_as_sf(grid_utm, coords = c("X", "Y"),
                           crs = sf::st_crs(paste0("+proj=utm +zone=", 
                                                    utm_zone, " +datum=WGS84")))
  grid_ll <- sf::st_transform(grid_sf, crs = 4326)
  coords_ll <- sf::st_coordinates(grid_ll)
  grid_utm$lon <- coords_ll[, 1]
  grid_utm$lat <- coords_ll[, 2]
} else {
  # Approximate back-conversion if sf not available
  cat("sf package not available; using approximate lon/lat from linear interpolation\n")
  grid_utm$lon <- approx(d$X, d$lon, xout = grid_utm$X)$y
  grid_utm$lat <- approx(d$Y, d$lat, xout = grid_utm$Y)$y
}

# Trim grid to approximate convex hull of sampling locations
# to avoid extrapolating far beyond the data
if (requireNamespace("sf", quietly = TRUE)) {
  hull <- sf::st_convex_hull(
    sf::st_union(
      sf::st_as_sf(d, coords = c("X", "Y"), crs = sf::st_crs(paste0("+proj=utm +zone=", utm_zone, " +datum=WGS84")))
    )
  )
  # Buffer slightly (5 km)
  hull_buf <- sf::st_buffer(hull, dist = 5000)
  grid_sf_trim <- sf::st_as_sf(grid_utm, coords = c("X", "Y"),
                                crs = sf::st_crs(paste0("+proj=utm +zone=", utm_zone, " +datum=WGS84")))
  inside <- sf::st_intersects(grid_sf_trim, hull_buf, sparse = FALSE)[, 1]
  grid_utm <- grid_utm[inside, ]
  cat("Grid cells after trimming to hull:", nrow(grid_utm), "\n")
}

```

Predict on the grid using m9 (spatial + family-specific spatial effects)
```{r}

m9 <- sdmTMB(
  pw ~ 1 + family_f,
  data = d,
  time_varying = ~1,
  time_varying_type = "ar1",
  mesh = mesh,
  extra_time = 2023,
  time = "year",
  family = tweedie(link = "log"),
  spatial = "off",
  spatial_varying = ~0 + family_f,
  spatiotemporal = "off"
)

sanity(m9)
```

Spatial predictions per family from m9
Predict on the regular grid for a representative year.

```{r}
# Create prediction data: replicate grid for each family
nd_spatial <- grid_utm |>
  mutate(year = 2020) |>   # representative year
  replicate_df(time_values = unique(d$family_f), time_name = "family_f") |>
  droplevels()

p_spatial <- predict(m9, newdata = nd_spatial, re_form_iid = NA)

# Plot spatially-varying coefficients (zeta) per family
zeta_cols <- grep("^zeta_s_", names(p_spatial), value = TRUE)

if (length(zeta_cols) > 0) {
  p_spatial |>
    pivot_longer(all_of(zeta_cols), names_to = "var", values_to = "zeta") |>
    # Clean up variable names for labels
    mutate(var = str_remove(var, "zeta_s_family_f")) |>
    ggplot(aes(lon, lat, color = zeta)) +
    geom_point(size = 0.5) +
    facet_wrap(~var) +
    scale_color_gradient2(
      low = "blue", mid = "white", high = "red",
      midpoint = 0, name = "Spatial\ncoefficient"
    ) +
    coord_sf() +
    labs(title = "Spatially-varying prey family coefficients (m9)",
         subtitle = "Positive = higher prey weight in that area",
         x = "Longitude", y = "Latitude")

  # Individual maps for key families
  for (fam in c("Gadidae", "Clupeidae")) {
    zeta_var <- paste0("zeta_s_family_f", fam)
    if (zeta_var %in% names(p_spatial)) {
      p_plot <- p_spatial |>
        filter(family_f == fam) |>
        distinct(lon, lat, .data[[zeta_var]])

      print(
        ggplot(p_plot, aes(lon, lat, color = .data[[zeta_var]])) +
          geom_point(size = 1.5) +
          scale_color_gradient2(
            low = "blue", mid = "white", high = "red",
            midpoint = 0, name = "Spatial\ncoefficient"
          ) +
          coord_sf() +
          labs(title = paste("Spatial coefficient —", fam, "(m9)"),
               x = "Longitude", y = "Latitude")
      )
    }
  }
}
```


Combined prediction map (total prey weight)

```{r}
# Predicted prey weight on response scale
p_spatial <- p_spatial |>
  mutate(pw_pred = exp(est))

# Total prey weight across all families
total_map <- p_spatial |>
  summarise(total_pw = sum(pw_pred), .by = c(lon, lat, X, Y))

ggplot(total_map, aes(lon, lat, color = total_pw)) +
  geom_point(size = 1) +
  scale_color_viridis_c(option = "plasma", name = "Total\nprey wt (g)") +
  coord_sf() +
  labs(title = "Total predicted prey weight — spatial map (m9, year 2020)",
       x = "Longitude", y = "Latitude") +
  # Overlay observed sampling locations
  geom_point(data = d |> distinct(lon, lat),
             aes(lon, lat), color = "grey30", shape = 3, size = 0.5,
             inherit.aes = FALSE)

# Per-family maps
p_spatial |>
  ggplot(aes(lon, lat, color = pw_pred)) +
  geom_point(size = 0.5) +
  facet_wrap(~family_f) +
  scale_color_viridis_c(option = "plasma", name = "Prey wt (g)",
                         trans = "log1p") +
  coord_sf() +
  labs(title = "Predicted prey weight by family — spatial map (m9)",
       x = "Longitude", y = "Latitude")

```

Re-fit m8 (simpler spatial-varying, no time-varying intercept)
and predict on the grid for comparison

```{r}
m8 <- sdmTMB(
  pw ~ 0 + family_f + year_f,
  data = d,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "off",
  spatial_varying = ~0 + family_f
)

sanity(m8)

nd_m8 <- grid_utm |>
  mutate(year_f = as.factor("2020")) |>
  replicate_df(time_values = unique(d$family_f), time_name = "family_f") |>
  droplevels()

p_m8 <- predict(m8, newdata = nd_m8, re_form_iid = NA) |>
  mutate(pw_pred = exp(est))

# Total prey weight map from m8
total_m8 <- p_m8 |>
  summarise(total_pw = sum(pw_pred), .by = c(lon, lat))

ggplot(total_m8, aes(lon, lat, color = total_pw)) +
  geom_point(size = 1) +
  scale_color_viridis_c(option = "plasma", name = "Total\nprey wt (g)") +
  coord_sf() +
  labs(title = "Total predicted prey weight — m8 (year 2020)",
       x = "Longitude", y = "Latitude")

```

Spatial maps over time: Multi-year spatial predictions from m8

```{r}
# Select representative years
key_years <- c(2012, 2016, 2020, 2024)
key_years <- key_years[key_years %in% unique(d$year)]

if (length(key_years) > 1) {
  nd_years <- grid_utm |>
    replicate_df(time_values = as.character(key_years), time_name = "year_f") |>
    mutate(year_f = as.factor(year_f)) |>
    replicate_df(time_values = unique(d$family_f), time_name = "family_f") |>
    droplevels()

  p_years <- predict(m8, newdata = nd_years, re_form_iid = NA) |>
    mutate(pw_pred = exp(est),
           year = as.numeric(as.character(year_f)))

  # Total prey weight by year
  total_years <- p_years |>
    summarise(total_pw = sum(pw_pred), .by = c(lon, lat, year))

  ggplot(total_years, aes(lon, lat, color = total_pw)) +
    geom_point(size = 0.5) +
    facet_wrap(~year) +
    scale_color_viridis_c(option = "plasma", name = "Total\nprey wt (g)") +
    coord_sf() +
    labs(title = "Total predicted prey weight across years (m8)",
         x = "Longitude", y = "Latitude")
}
```

Follow up on:
2. Can we estimate a spatiotemporal random field? (different spatial predictions every year?)

3. s(day, by = prey?)? This would allow for seasonal variation in prey!

