---
title: "Harbour Seal Diet Hybrid Prey-Group Models"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    page-layout: full
    embed-resources: true
editor: source
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
---

## Purpose

This notebook builds a hybrid prey-group modelling pipeline for the harbour seal diet data.

The hybrid approach uses species-level resolution for Gadidae (because of their ecological and policy relevance) while keeping all other prey at the family level used in `03b-diet-models-family.qmd`.

Specifically:

-   Gadidae are split into: Gadus morhua, Merlangius merlangus, Pollachius sp. / Melanogrammus aeglefinus, unidentified Gadidae, and Other Gadidae.
-   Clupeidae, Pleuronectidae, Gobiidae, and Other remain as grouped families.

This gives 9 prey groups instead of the 5 families used in `03b-diet-models-family.qmd`.

See `decision.md` at the project root for the rationale behind this design choice.

The workflow is:

-   retain colonies through Onsala only,
-   retain Faeces and Hunt samples only,
-   build a sample-by-prey-group modelling table with zero expansion,
-   use 2013-2025 as the main analysis period,
-   model prey biomass by prey group with collection_method and group-specific smooth year effects

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(ggplot2)
library(forcats)
library(stringr)
library(janitor)
library(sf)
library(sdmTMB)
library(purrr)
library(rnaturalearth)
library(sf)
library(ggsidekick)

theme_set(theme_sleek())

home <- here::here()

biomass_file <- file.path(home, "Data", "HS_SKAG_estimated_abundance_biomass_130426.csv")
ices_areas_file <- file.path(home, "Data", "shapefiles", "ICES_areas", "ICES_Areas_20160601_cut_dense_3857.shp")

colony_keep <- c("Koster", "Väderöarna", "Lysekil", "Marstrand", "Onsala")
method_keep <- c("Faeces", "Hunt")
analysis_year_min <- 2013L

# Gadidae species to keep at species level
gad_species_keep <- c(
  "Gadus morhua",
  "Merlangius merlangus",
  "Pollachius sp. / Melanogrammus aeglefinus",
  "Gadidae"
)

# All prey groups in the hybrid resolution
prey_group_levels <- c(
  gad_species_keep,
  "Other Gadidae",
  "Clupeidae",
  "Pleuronectidae",
  "Gobiidae",
  "Other"
)

prey_group_palette <- c(
  "Gadus morhua"                              = "#2E6B1F",
  "Merlangius merlangus"                      = "#5DAE4A",
  "Pollachius sp. / Melanogrammus aeglefinus"  = "#A3D977",
  "Gadidae"                                   = "#3B8C27",
  "Other Gadidae"                             = "#7EC87E",
  "Clupeidae"                                 = "#1F5AA6",
  "Pleuronectidae"                            = "#D94FD3",
  "Gobiidae"                                  = "#F39C12",
  "Other"                                     = "#9E4A7A",
  "Total"                                     = "#000000"
)

normalize_colony_name <- function(x) {
  x <- trimws(as.character(x))
  case_when(
    grepl("der.*arna", x, ignore.case = TRUE) ~ "Väderöarna",
    grepl("^Hallands.*der", x, ignore.case = TRUE) ~ "Hallands Väderö",
    TRUE ~ x
  )
}
```

## Read and harmonize source data

```{r read-data}
sample_meta_workbook <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year_wb = suppressWarnings(as.integer(year)),
    month_wb = suppressWarnings(as.integer(month)),
    quarter_wb = suppressWarnings(as.integer(quarter)),
    latitude_wb = suppressWarnings(as.numeric(latitude)),
    longitude_wb = suppressWarnings(as.numeric(longitude)),
    colony_wb = normalize_colony_name(colony),
    collection_method_wb = if_else(
      is.na(collection_method) | collection_method == "",
      "Unknown",
      as.character(collection_method)
    )
  )

diet_raw <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  left_join(sample_meta_workbook, by = "sample_id") |>
  mutate(
    year = coalesce(year_wb, suppressWarnings(as.integer(year))),
    month = coalesce(month_wb, suppressWarnings(as.integer(month))),
    quarter = coalesce(quarter_wb, suppressWarnings(as.integer(quarter))),
    latitude = coalesce(latitude_wb, suppressWarnings(as.numeric(latitude))),
    longitude = coalesce(longitude_wb, suppressWarnings(as.numeric(longitude))),
    colony = coalesce(colony_wb, normalize_colony_name(colony)),
    collection_method = coalesce(collection_method_wb, as.character(collection_method)),
    collection_method = if_else(is.na(collection_method) | collection_method == "", "Unknown", collection_method),
    colony = if_else(is.na(colony) | colony == "", "Unknown", colony),
    family = if_else(is.na(family) | family == "", "Unknown", family),
    prey_species = prey_species_lat,
    quarter = coalesce(quarter, ((month - 1L) %/% 3L) + 1L)
  ) |>
  select(-ends_with("_wb"))

sample_meta_all <- diet_raw |>
  filter(!is.na(sample_id), !is.na(year), !is.na(latitude), !is.na(longitude)) |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year,
    month,
    quarter,
    lat = latitude,
    lon = longitude,
    collection_method,
    colony
  )

# --- Hybrid prey-group assignment ---
# Gadidae split to species; all other families kept at family level.
diet_rows_all <- diet_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species),
    prey_species != "",
    tolower(prey_species) != "non prey item"
  ) |>
  transmute(
    sample_id,
    prey_group = case_when(
      family == "Gadidae" & prey_species %in% gad_species_keep ~ prey_species,
      family == "Gadidae" ~ "Other Gadidae",
      family == "Clupeidae" ~ "Clupeidae",
      family == "Pleuronectidae" ~ "Pleuronectidae",
      family == "Gobiidae" ~ "Gobiidae",
      TRUE ~ "Other"
    ),
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, prey_group))

sample_totals_all <- diet_rows_all |>
  summarise(sum_biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = sample_id)

sample_meta_filtered <- sample_meta_all |>
  left_join(sample_totals_all, by = "sample_id") |>
  mutate(sum_biomass_kg = coalesce(sum_biomass_kg, 0)) |>
  filter(
    collection_method %in% method_keep,
    colony %in% colony_keep,
    sum_biomass_kg > 0
  )

if (nrow(sample_meta_filtered) == 0) {
  stop("No samples remain after filtering by colony, method, and positive total biomass.")
}

analysis_year_max <- max(sample_meta_filtered$year, na.rm = TRUE)

sample_meta_main <- sample_meta_filtered |>
  filter(year >= analysis_year_min, year <= analysis_year_max)

diet_rows_main <- diet_rows_all |>
  semi_join(sample_meta_main, by = "sample_id")
```

## Support summaries and prey-group grouping

```{r support-tables}
n_main_samples <- nrow(sample_meta_main)

prey_group_support_main <- diet_rows_main |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = prey_group
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence,
    prey_group = factor(prey_group, levels = prey_group_levels)
  ) |>
  arrange(prey_group)
```

```{r build-model-table}
year_center <- mean(sample_meta_main$year, na.rm = TRUE)

diet_model <- sample_meta_main |>
  select(sample_id, year, month, quarter, lat, lon, collection_method, colony, sum_biomass_kg) |>
  tidyr::expand_grid(prey_group = prey_group_levels) |>
  left_join(diet_rows_main, by = c("sample_id", "prey_group")) |>
  mutate(
    biomass_kg = coalesce(biomass_kg, 0),
    present = as.integer(biomass_kg > 0),
    prop_biomass = if_else(sum_biomass_kg > 0, biomass_kg / sum_biomass_kg, 0),
    year_ct = year - year_center,
    collection_method = factor(collection_method, levels = method_keep),
    colony = factor(colony, levels = colony_keep),
    prey_group = factor(prey_group, levels = prey_group_levels),
    year_f = as.factor(year),
    quarter_f = factor(quarter),
    month_f = factor(month)
  ) |>
  add_utm_columns(ll_names = c("lon", "lat"))

diet_model |>
  summarise(
    n_samples = n_distinct(sample_id),
    n_rows = n(),
    min_year = min(year),
    max_year = max(year),
    n_prey_groups = n_distinct(prey_group)
  )
```

## Support diagnostics

```{r support-plots}
samples_by_year_method <- sample_meta_main |>
  count(year, collection_method, name = "n_samples")

samples_by_colony_method <- sample_meta_main |>
  count(colony, collection_method, name = "n_samples")

sample_map <- sample_meta_main |>
  ggplot(aes(lon, lat, colour = collection_method)) +
  geom_point(alpha = 0.8, size = 2) +
  coord_sf() +
  labs(
    title = "Retained samples in the main analysis period",
    subtitle = "Colonies through Onsala only; methods restricted to Faeces and Hunt",
    x = "Longitude",
    y = "Latitude",
    colour = "Method"
  )

plot_year_method <- samples_by_year_method |>
  ggplot(aes(year, n_samples, fill = collection_method)) +
  geom_col(position = "stack") +
  labs(
    title = "Samples by year and collection method",
    x = "Year",
    y = "Samples",
    fill = "Method"
  )

plot_colony_method <- samples_by_colony_method |>
  ggplot(aes(colony, n_samples, fill = collection_method)) +
  geom_col(position = "stack") +
  labs(
    title = "Samples by colony and collection method",
    subtitle = "Colony is retained for support diagnostics only",
    x = NULL,
    y = "Samples",
    fill = "Method"
  )

plot_prey_group_support <- prey_group_support_main |>
  mutate(prey_group = fct_reorder(as.character(prey_group), total_biomass_kg)) |>
  ggplot(aes(total_biomass_kg, prey_group)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Hybrid prey groups retained for modelling",
    subtitle = "Gadidae split to species; other families grouped",
    x = "Total biomass (kg)",
    y = NULL
  )

plot_zero_rate <- prey_group_support_main |>
  mutate(prey_group = fct_reorder(as.character(prey_group), zero_rate)) |>
  ggplot(aes(zero_rate, prey_group)) +
  geom_col(fill = "tomato") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Zero proportion by prey group",
    x = "Zero proportion",
    y = NULL
  )

sample_map
plot_year_method
plot_colony_method
plot_prey_group_support
plot_zero_rate
```



```{r}
# Gridded map prediction
d <- diet_model

x <- d$X
y <- d$Y
z <- chull(x, y)

coords <- cbind(x[z], y[z])

coords <- rbind(coords, coords[1, ])

plot(coords[, 1] ~ coords[, 2]) # plot data

sp_poly <- sp::SpatialPolygons(
  list(sp::Polygons(list(sp::Polygon(coords)), ID = 1))
)

sp_poly_df <- sp::SpatialPolygonsDataFrame(sp_poly,
  data = data.frame(ID = 1)
)
cell_width <- 0.5

pred_grid <- expand.grid(
  X = seq(min(d$X), max(d$X), cell_width),
  Y = seq(min(d$Y), max(d$Y), cell_width),
  year = c(min(d$year):max(d$year))
)

sp::coordinates(pred_grid) <- c("X", "Y")

inside <- !is.na(sp::over(pred_grid, as(sp_poly_df, "SpatialPolygons")))

pred_grid <- pred_grid[inside, ]

pred_grid <- as.data.frame(pred_grid) |> 
  mutate(collection_method = "Hunt",
         year_ct = year - year_center,
         year_f = as.factor(year))

pred_grid <- replicate_df(pred_grid,
    time_name = "prey_group",
    time_values = unique(d$prey_group)
    )
```

## Map plot

```{r}
# Add coastline
land <- ne_countries(scale = "large", returnclass = "sf") |>
  st_transform(32632)
```

## Mesh

```{r mesh}
mesh_cutoff_km <- 0.5

mesh <- make_mesh(diet_model, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

plot(mesh)
```

## Candidate models

```{r fit-models}
#| cache: true
# Spatial no time
m1 <- sdmTMB(
  formula = biomass_kg ~ 0 + prey_group + collection_method,
  data = diet_model,
  family = tweedie(),
  mesh = mesh,
  spatial = "on"
)

# smooth year by prey_group
m2 <- sdmTMB(
  formula = biomass_kg ~ 0 + prey_group + collection_method + s(year_ct, by = prey_group),
  data = diet_model,
  family = tweedie(),
  mesh = mesh,
  spatial = "on"
)

# SVC prey_group + year smooth
m3 <- sdmTMB(
  formula = biomass_kg ~ 0 + prey_group + collection_method + s(year_ct, by = prey_group),
  data = diet_model,
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  spatial_varying = ~0 + prey_group
)

# SVC prey_group + year smooth + iid
m4 <- sdmTMB(
  formula = biomass_kg ~ 0 + prey_group + collection_method + s(year_ct, by = prey_group),
  data = diet_model,
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  time = "year",
  spatiotemporal = "iid",
  spatial_varying = ~0 + prey_group
)

# factor year by prey_group, collection by prey_group, no st
m5 <- sdmTMB(
  formula = biomass_kg ~ 0 + prey_group + collection_method + collection_method:prey_group + year_f:prey_group,
  data = diet_model,
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  time = "year",
  spatiotemporal = "off",
  spatial_varying = ~0 + prey_group
)

# factor year by prey_group, collection by prey_group, iid st
m6 <- sdmTMB(
  formula = biomass_kg ~ 0 + prey_group + collection_method + collection_method:prey_group + year_f:prey_group,
  data = diet_model,
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  time = "year",
  spatiotemporal = "iid",
  spatial_varying = ~0 + prey_group
)

# SVC prey_group, random walk time by prey_group
AIC(m1, m2, m3, m4, m5, m6) |> arrange(AIC)
```

```{r}
m6
sanity(m6)

# Predict over time
ndt <- tibble(
  expand_grid(
    year = unique(diet_model$year),
    prey_group = unique(diet_model$prey_group)))  |> 
  mutate(collection_method = "Hunt",
         year_f = as.factor(year),
         year = round(year))
  
p <- predict(m6, newdata = ndt, re_form = NA, se_fit = TRUE)

tot_pred <- predict(m6, newdata = ndt, re_form = NA, nsim = 500)

tot_pred <- tot_pred |> 
  as.data.frame() |> 
  bind_cols(ndt) |> 
  pivot_longer(cols = starts_with("V"), names_to = "sim") 

tot_pred |> 
  filter(sim == "V1" & year == 2024)

tot <- tot_pred |> 
  mutate(value = exp(value)) |> 
  summarise(sum = sum(value), .by = c(year, sim)) |> 
  summarise(est = median(sum),
            lwr = quantile(sum, probs = 0.05),
            upr = quantile(sum, probs = 0.95), 
            sd = sd(sum),
            .by = year) |> 
  mutate(prey_group = "Total")

p |> 
  mutate(lwr = exp(est - 1.96*est_se),
         upr = exp(est + 1.96*est_se),
         est = exp(est)) |> 
  bind_rows(tot) |> 
  ggplot(aes(year, est, color = prey_group, fill = prey_group)) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr),
              alpha = 0.3, color = NA) +
  scale_x_continuous(breaks = seq(2012, 2025, by = 3)) +
  scale_color_manual(values = prey_group_palette) +
  scale_fill_manual(values = prey_group_palette) +
  theme(legend.position = "right") +
  guides(fill = "none") +
  labs(y = "Kg/sample", color = "Prey group") +
  facet_wrap(~prey_group, scales = "free") +
  geom_line()

# Significant increase in total prey? - NO!
# Use 1/sd as weights in linear regression!
mod <- lm(est ~ year,
          weights = 1/sd,
          data = tot)

summary(mod)

# Plot proportion over time
p |> 
  mutate(sum = sum(exp(est)), 
         prop = exp(est)/sum,
         .by = c(year)) |> 
  ggplot(aes(year, prop, color = prey_group, fill = prey_group)) + 
  scale_x_continuous(breaks = seq(2012, 2025, by = 3)) +
  scale_color_manual(values = prey_group_palette) +
  scale_fill_manual(values = prey_group_palette) +
  coord_cartesian(expand = 0) +
  geom_area()
```

```{r}
# Predict!
pg <- predict(m6, newdata = pred_grid)

# Plot SVCs
pg |> 
  filter(year == 2013) |> 
  pivot_longer(starts_with("zeta_s_prey_group"),
               names_to = "var") |> 
  mutate(var = str_remove(var, "zeta_s_prey_group")) |> 
  ggplot(aes(X*1000, Y*1000, fill = value)) +
  geom_raster() +
  facet_wrap(~var, ncol = 5) +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
    coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2()

# Plot spatial predictions in a given year
pg |> 
  filter(year == 2013) |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~prey_group, ncol = 5)

# Now do by year for cod
pg |> 
  filter(prey_group == "Gadus morhua") |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  guides(fill = guide_colorbar(position = "inside")) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position.inside = c(0.95, 0.2)) +
  facet_wrap(~year, ncol = 7)

# ... and whiting
pg |> 
  filter(prey_group == "Merlangius merlangus") |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  guides(fill = guide_colorbar(position = "inside")) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position.inside = c(0.95, 0.2)) +
  facet_wrap(~year, ncol = 7)

# ... and clupeids (kept at family level)
pg |> 
  filter(prey_group == "Clupeidae") |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  guides(fill = guide_colorbar(position = "inside")) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position.inside = c(0.95, 0.2)) +
  facet_wrap(~year, ncol = 7)
```

Plot effect of collection method:

```{r}
nd_col <- tibble(
  expand_grid(
    collection_method = unique(diet_model$collection_method),
    prey_group = unique(diet_model$prey_group)
  )) |> 
  mutate(year = 2013,
         year_f = as.factor(year))

pcol <- predict(m6, newdata = nd_col, se_fit = TRUE, re_form = NA)

ggplot(pcol, aes(exp(est), prey_group, color = collection_method)) + 
  geom_point(position = position_dodge(width = 0.2)) + 
  geom_errorbar(aes(xmin = exp(est - 1.96*est_se),
                    xmax = exp(est + 1.96*est_se)),
                width = 0,
                position = position_dodge(width = 0.2))
```




