---
title: "Harbour Seal Ontogenetic Diet Family Models"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    page-layout: full
    embed-resources: true
editor: source
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
---

## Purpose

This notebook mirrors `03b-diet-models-family.qmd`, but focuses on ontogenetic variation in diet composition.

The workflow is:

- retain colonies through Onsala only,
- retain hunted seals only,
- keep the same dominant prey-family grouping as `03b`,
- build a hunted-seal sample-by-family modelling table,
- test whether seal length changes diet composition,
- then test whether sex adds further structure once length is included.

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(forcats)
library(stringr)
library(janitor)
library(sf)
library(sdmTMB)
library(purrr)
library(rnaturalearth)
library(ggsidekick)

theme_set(theme_sleek())

home <- here::here()

biomass_file <- file.path(home, "Data", "HS_SKAG_estimated_abundance_biomass_130426.csv")

colony_keep <- c("Koster", "Vaderoarna", "Lysekil", "Marstrand", "Onsala")
analysis_year_min <- 2013L
dominant_families <- c("Gadidae", "Clupeidae", "Pleuronectidae", "Gobiidae")

family_palette <- c(
  "Gadidae"        = "#2E6B1F",
  "Clupeidae"      = "#1F5AA6",
  "Pleuronectidae" = "#D94FD3",
  "Gobiidae"       = "#F39C12",
  "Other"          = "#9E4A7A",
  "Total"          = "#000000"
)

normalize_colony_name <- function(x) {
  x <- trimws(as.character(x))
  case_when(
    grepl("der.*arna", x, ignore.case = TRUE) ~ "Vaderoarna",
    grepl("^Hallands.*der", x, ignore.case = TRUE) ~ "Hallands Vadero",
    TRUE ~ x
  )
}
```

## Read and harmonize source data

```{r read-data}
sample_meta_workbook <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year_wb = suppressWarnings(as.integer(year)),
    month_wb = suppressWarnings(as.integer(month)),
    quarter_wb = suppressWarnings(as.integer(quarter)),
    latitude_wb = suppressWarnings(as.numeric(latitude)),
    longitude_wb = suppressWarnings(as.numeric(longitude)),
    colony_wb = normalize_colony_name(colony),
    sex_wb = case_when(
      sex %in% c("Female", "Male") ~ as.character(sex),
      TRUE ~ NA_character_
    ),
    age_wb = suppressWarnings(as.numeric(age)),
    body_length_cm_wb = suppressWarnings(as.numeric(body_length_cm)),
    collection_method_wb = if_else(
      is.na(collection_method) | collection_method == "",
      "Unknown",
      as.character(collection_method)
    )
  )

diet_raw <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  left_join(sample_meta_workbook, by = "sample_id") |>
  mutate(
    year = coalesce(year_wb, suppressWarnings(as.integer(year))),
    month = coalesce(month_wb, suppressWarnings(as.integer(month))),
    quarter = coalesce(quarter_wb, suppressWarnings(as.integer(quarter))),
    latitude = coalesce(latitude_wb, suppressWarnings(as.numeric(latitude))),
    longitude = coalesce(longitude_wb, suppressWarnings(as.numeric(longitude))),
    colony = coalesce(colony_wb, normalize_colony_name(colony)),
    collection_method = coalesce(collection_method_wb, as.character(collection_method)),
    collection_method = if_else(is.na(collection_method) | collection_method == "", "Unknown", collection_method),
    family = if_else(is.na(family) | family == "", "Unknown", family),
    prey_species = prey_species_lat,
    sex = coalesce(sex_wb, case_when(
      sex %in% c("Female", "Male") ~ as.character(sex),
      TRUE ~ NA_character_
    )),
    age = coalesce(age_wb, suppressWarnings(as.numeric(age))),
    body_length_cm = coalesce(body_length_cm_wb, suppressWarnings(as.numeric(body_length_cm))),
    quarter = coalesce(quarter, ((month - 1L) %/% 3L) + 1L)
  ) |>
  select(-ends_with("_wb"))

sample_meta_all <- diet_raw |>
  filter(!is.na(sample_id), !is.na(year), !is.na(latitude), !is.na(longitude)) |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year,
    month,
    quarter,
    lat = latitude,
    lon = longitude,
    collection_method,
    colony,
    sex,
    age,
    body_length_cm
  )

diet_rows_all <- diet_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species),
    prey_species != "",
    tolower(prey_species) != "non prey item"
  ) |>
  transmute(
    sample_id,
    family,
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, family))

sample_totals_all <- diet_rows_all |>
  summarise(sum_biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = sample_id)

sample_meta_filtered <- sample_meta_all |>
  left_join(sample_totals_all, by = "sample_id") |>
  mutate(sum_biomass_kg = coalesce(sum_biomass_kg, 0)) |>
  filter(
    collection_method == "Hunt",
    colony %in% colony_keep,
    sum_biomass_kg > 0
  )

if (nrow(sample_meta_filtered) == 0) {
  stop("No hunted samples remain after filtering by colony and positive total biomass.")
}

analysis_year_max <- max(sample_meta_filtered$year, na.rm = TRUE)

sample_meta_main <- sample_meta_filtered |>
  filter(year >= analysis_year_min, year <= analysis_year_max)

diet_rows_main <- diet_rows_all |>
  semi_join(sample_meta_main, by = "sample_id")
```

## Support summaries and family grouping

```{r support-tables}
n_main_samples <- nrow(sample_meta_main)

family_levels <- unique(c(dominant_families, "Other"))

family_support_main <- diet_rows_main |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = family
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence
  ) |>
  arrange(desc(total_biomass_kg))

diet_rows_grouped <- diet_rows_main |>
  mutate(family = if_else(family %in% dominant_families, family, "Other")) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, family))

family_support_grouped <- diet_rows_grouped |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = family
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence,
    family = factor(family, levels = family_levels)
  ) |>
  arrange(family)
```

## Build hunted-only model table

```{r build-model-table}
year_center <- mean(sample_meta_main$year, na.rm = TRUE)

diet_model_all <- sample_meta_main |>
  select(sample_id, year, month, quarter, lat, lon, colony, sum_biomass_kg, sex, age, body_length_cm) |>
  expand_grid(family = family_levels) |>
  left_join(diet_rows_grouped, by = c("sample_id", "family")) |>
  mutate(
    biomass_kg = coalesce(biomass_kg, 0),
    present = as.integer(biomass_kg > 0),
    prop_biomass = if_else(sum_biomass_kg > 0, biomass_kg / sum_biomass_kg, 0),
    year_ct = year - year_center,
    colony = factor(colony, levels = colony_keep),
    family = factor(family, levels = family_levels),
    sex = factor(sex, levels = c("Female", "Male")),
    year_f = factor(year),
    quarter_f = factor(quarter),
    month_f = factor(month)
  ) |>
  add_utm_columns(ll_names = c("lon", "lat"))

diet_model <- diet_model_all |>
  filter(!is.na(body_length_cm))

diet_model_sex <- diet_model |>
  filter(!is.na(sex))

diet_model |>
  summarise(
    n_samples = n_distinct(sample_id),
    n_rows = n(),
    min_year = min(year),
    max_year = max(year),
    n_families = n_distinct(family),
    length_min = min(body_length_cm),
    length_max = max(body_length_cm)
  )
```

## Support diagnostics

```{r support-plots}
hunt_samples_by_year <- sample_meta_main |>
  count(year, name = "n_samples")

missing_covariates <- sample_meta_main |>
  summarise(
    n_samples = n(),
    n_length = sum(!is.na(body_length_cm)),
    n_sex = sum(!is.na(sex)),
    n_age = sum(!is.na(age)),
    n_length_sex = sum(!is.na(body_length_cm) & !is.na(sex)),
    n_all_three = sum(!is.na(body_length_cm) & !is.na(sex) & !is.na(age))
  )

age_length_plot <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(age), !is.na(sex)) |>
  ggplot(aes(age, body_length_cm, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    title = "Relation between seal age and length",
    subtitle = "Separate fitted lines for female and male hunted seals",
    x = "Age",
    y = "Body length (cm)",
    color = "Sex"
  )

length_histogram <- sample_meta_main |>
  filter(!is.na(body_length_cm)) |>
  ggplot(aes(body_length_cm)) +
  geom_histogram(binwidth = 5, color = "white", fill = "grey40") +
  labs(
    title = "Seal length distribution in hunted samples",
    x = "Body length (cm)",
    y = "Samples"
  )

length_histogram_sex_colony <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(sex)) |>
  mutate(colony = factor(colony, levels = colony_keep)) |>
  ggplot(aes(body_length_cm, fill = sex)) +
  geom_histogram(binwidth = 5, color = "white") +
  facet_wrap(~colony, ncol = 1, scales = "free_y") +
  labs(
    title = "Seal length distribution in hunted samples by colony and sex",
    subtitle = "Colonies ordered from north to south",
    x = "Body length (cm)",
    y = "Samples",
    fill = "Sex"
  )

length_by_year <- sample_meta_main |>
  filter(!is.na(body_length_cm)) |>
  ggplot(aes(year, body_length_cm)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Seal length over time",
    x = "Year",
    y = "Body length (cm)"
  )

sex_balance <- sample_meta_main |>
  filter(!is.na(sex)) |>
  count(sex) |>
  ggplot(aes(sex, n, fill = sex)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Sex balance in hunted samples",
    x = NULL,
    y = "Samples"
  )

sex_balance_colony <- sample_meta_main |>
  filter(!is.na(sex)) |>
  mutate(colony = factor(colony, levels = colony_keep)) |>
  count(colony, sex, name = "n") |>
  group_by(colony) |>
  mutate(
    total_n = sum(n),
    prop = n / total_n
  ) |>
  ungroup() |>
  ggplot(aes(colony, prop, fill = sex)) +
  geom_col() +
  geom_text(
    data = ~ distinct(.x, colony, total_n),
    aes(x = colony, y = 1.02, label = total_n),
    inherit.aes = FALSE,
    vjust = 0
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1.08),
    expand = c(0, 0)
  ) +
  labs(
    title = "Relative sex composition by colony",
    subtitle = "Numbers above bars show total hunted samples with known sex",
    x = NULL,
    y = "Relative abundance",
    fill = "Sex"
  )

family_support_plot <- family_support_grouped |>
  mutate(family = fct_reorder(as.character(family), total_biomass_kg)) |>
  ggplot(aes(total_biomass_kg, family)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Grouped prey families retained for modelling",
    subtitle = "Same dominant-family set as 03b",
    x = "Total biomass (kg)",
    y = NULL
  )

missing_covariates
hunt_samples_by_year
age_length_plot
length_histogram
length_histogram_sex_colony
length_by_year
sex_balance
sex_balance_colony
family_support_plot
```

## Prediction grid

```{r prediction-grid}
d <- diet_model

hull_idx <- chull(d$X, d$Y)
hull_coords <- cbind(d$X[hull_idx], d$Y[hull_idx])
hull_coords <- rbind(hull_coords, hull_coords[1, ])

sp_poly <- sp::SpatialPolygons(
  list(sp::Polygons(list(sp::Polygon(hull_coords)), ID = 1))
)

sp_poly_df <- sp::SpatialPolygonsDataFrame(
  sp_poly,
  data = data.frame(ID = 1)
)

cell_width <- 0.5

pred_grid_base <- expand.grid(
  X = seq(min(d$X), max(d$X), cell_width),
  Y = seq(min(d$Y), max(d$Y), cell_width),
  year = seq(min(d$year), max(d$year))
)

sp::coordinates(pred_grid_base) <- c("X", "Y")

inside <- !is.na(sp::over(pred_grid_base, as(sp_poly_df, "SpatialPolygons")))

pred_grid_base <- pred_grid_base[inside, ]

pred_grid_base <- as.data.frame(pred_grid_base) |>
  mutate(
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model$year_f))
  )

land <- ne_countries(scale = "large", returnclass = "sf") |>
  st_transform(32632)
```

## Mesh

```{r mesh}
mesh_cutoff_km <- 0.5

mesh <- make_mesh(diet_model, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

plot(mesh)
```


## Shared prediction reference

```{r shared-prediction-reference}
prediction_reference <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(sex)) |>
  summarise(
    pred_year = round(median(sort(unique(year)))),
    pred_length_cm = median(body_length_cm, na.rm = TRUE)
  )

pred_year_common <- prediction_reference$pred_year
pred_length_common <- prediction_reference$pred_length_cm
```


## Family-level Gadidae questions: sex, length, year, space

This section keeps the earlier family-level workflow. It fits one joint family model on all retained family groups, converts predictions to diet composition, and then filters the prediction output to `Gadidae`.

```{r fit-gadidae-models}
#| cache: true
diet_model_gad <- diet_model_sex |>
  mutate(body_length_ct = body_length_cm - mean(body_length_cm, na.rm = TRUE))

mesh_gad <- make_mesh(diet_model_gad, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

# Do females and males differ in diet choice? After accounting for length
g1 <- sdmTMB(
  formula = biomass_kg ~ sex*family + body_length_cm + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "on"
)

# Length-specific diet
g2 <- sdmTMB(
  formula = biomass_kg ~ sex*family + body_length_cm*family + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "on"
)

# Length-specific diet (nonlinear ontogenetic diet change)
g3 <- sdmTMB(
  formula = biomass_kg ~ sex*family + s(body_length_cm, by = family) + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "on"
)

# Different length effects by sex
g4 <- sdmTMB(
  formula = biomass_kg ~ sex*family + body_length_cm*sex + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "on"
)

# Spatial fields by family
g5 <- sdmTMB(
  formula = biomass_kg ~ sex*family + body_length_cm + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + family
)

# Spatial fields by family with body length by family
g6 <- sdmTMB(
  formula = biomass_kg ~ sex*family + body_length_cm*family + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + family
)

# Three-way interaction
g7 <- sdmTMB(
  formula = biomass_kg ~ body_length_cm*family*sex + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + family
)

# g6 but with spatiotemporal fields
g8 <- sdmTMB(
  formula = biomass_kg ~ sex*family + body_length_cm*family + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "off",
  spatiotemporal = "iid",
  time = "year",
  spatial_varying = ~ 0 + family
)

# Does the effect of sex differ between quarters?
g9 <- sdmTMB(
  formula = biomass_kg ~ sex*family + quarter_f*sex + body_length_cm*family + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + family
)

# Simple seasonal effect
g10 <- sdmTMB(
  formula = biomass_kg ~ sex*family + quarter_f + body_length_cm*family + (1|year_f),
  data = diet_model_gad,
  mesh = mesh_gad,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + family
)

AIC(g1, g2, g3, g4, g5, g6, g7, g8, g9, g10) |>
  arrange(AIC)
```

```{r family-level-diagnostics}
# Season effects from the family-level model using the shared reference seal.
nd <- expand_grid(
  quarter_f = as.factor(c(1:4)),
  family = unique(diet_model_gad$family),
  sex = unique(diet_model_gad$sex)
) |>
  mutate(
    body_length_cm = pred_length_common,
    year = pred_year_common,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_gad$year_f)),
    X = median(diet_model_gad$X, na.rm = TRUE),
    Y = median(diet_model_gad$Y, na.rm = TRUE)
  )

p <- predict(g9, newdata = nd, re_form = NA, re_form_iid = NA)

ggplot(p, aes(sex, exp(est), fill = family)) +
  facet_wrap(~quarter_f, ncol = 4) +
  geom_bar(stat = "identity")

ggplot(p, aes(as.numeric(as.character(quarter_f)), exp(est), fill = family)) +
  facet_wrap(~sex, ncol = 4) +
  geom_bar(stat = "identity")

qqnorm(residuals(g9)); qqline(residuals(g9))

# Quarter effect for Gadidae by sex at the same reference seal and year.
nd <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_gad$quarter_f)),
  sex = unique(diet_model_gad$sex)
) |>
  mutate(
    body_length_cm = pred_length_common,
    family = "Gadidae",
    year = pred_year_common,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_gad$year_f)),
    X = median(diet_model_gad$X, na.rm = TRUE),
    Y = median(diet_model_gad$Y, na.rm = TRUE)
  )

p <- predict(g9, newdata = nd, re_form = NA, re_form_iid = NA)

ggplot(p, aes(as.numeric(as.character(quarter_f)), exp(est), color = sex)) +
  geom_line()

plot(ggeffects::ggpredict(g9, c("family", "sex", "quarter_f"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot(ggeffects::ggpredict(g6, c("family", "sex"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot(ggeffects::ggpredict(g6, c("body_length_cm", "family")))
plot(ggeffects::ggpredict(g6, c("body_length_cm", "family", "sex")))

diet_model_gad |>
  distinct(sample_id, body_length_cm, sex) |>
  ggplot(aes(body_length_cm, fill = sex)) +
  geom_histogram()

diet_model_gad |>
  distinct(sample_id, month, sex) |>
  ggplot(aes(as.factor(month), fill = sex)) +
  geom_bar()
```


## Hybrid prey-group questions: sex, length, year, space

This section uses one hybrid prey grouping for the hunted samples:

- split `Gadidae` into the most supported scientific-name labels,
- keep the remaining gadids as `Other Gadidae`,
- keep non-gadid prey at the same grouped family level used above.

```{r hybrid-prey-data}
gad_species_keep <- c(
  "Gadus morhua",
  "Merlangius merlangus",
  "Pollachius sp. / Melanogrammus aeglefinus",
  "Gadidae"
)

prey_group_levels <- c(
  gad_species_keep,
  "Other Gadidae",
  "Clupeidae",
  "Pleuronectidae",
  "Gobiidae",
  "Other"
)

hybrid_rows <- diet_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species),
    prey_species != "",
    tolower(prey_species) != "non prey item"
  ) |>
  transmute(
    sample_id,
    prey_group = case_when(
      family == "Gadidae" & prey_species %in% gad_species_keep ~ prey_species,
      family == "Gadidae" ~ "Other Gadidae",
      family == "Clupeidae" ~ "Clupeidae",
      family == "Pleuronectidae" ~ "Pleuronectidae",
      family == "Gobiidae" ~ "Gobiidae",
      TRUE ~ "Other"
    ),
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, prey_group))

prey_group_support <- hybrid_rows |>
  semi_join(sample_meta_main |> filter(!is.na(body_length_cm), !is.na(sex)), by = "sample_id") |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = prey_group
  ) |>
  mutate(prey_group = factor(prey_group, levels = prey_group_levels)) |>
  arrange(prey_group)

diet_model_hybrid <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(sex)) |>
  select(sample_id, year, month, quarter, lat, lon, colony, sum_biomass_kg, sex, body_length_cm) |>
  expand_grid(prey_group = prey_group_levels) |>
  left_join(hybrid_rows, by = c("sample_id", "prey_group")) |>
  mutate(
    biomass_kg = coalesce(biomass_kg, 0),
    year_ct = year - year_center,
    year_f = factor(year),
    quarter_f = factor(quarter),
    prey_group = factor(prey_group, levels = prey_group_levels),
    sex = factor(sex, levels = c("Female", "Male"))
  ) |>
  add_utm_columns(ll_names = c("lon", "lat"))

prey_group_support
```

```{r fit-hybrid-models}
#| cache: true
mesh_hybrid <- make_mesh(diet_model_hybrid, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

h1 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h2 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h3 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + s(body_length_cm, by = prey_group) + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h4 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm * sex + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h5 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h6 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h7 <- sdmTMB(
  formula = biomass_kg ~ body_length_cm * prey_group * sex + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h8 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatiotemporal = "iid",
  time = "year",
  spatial_varying = ~ 0 + prey_group
)

h9 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + quarter_f * sex + body_length_cm * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h10 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + quarter_f + body_length_cm * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

AIC(h1, h2, h3, h4, h5, h6, h7, h8, h9, h10) |>
  arrange(AIC)
```

```{r hybrid-level-diagnostics}
pred_year_hybrid <- pred_year_common
pred_length_hybrid <- pred_length_common

# Season effects from the hybrid model using the same reference seal and year as the family plots.
nd_hybrid <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels),
  sex = unique(diet_model_hybrid$sex)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_hybrid <- predict(h9, newdata = nd_hybrid, re_form = NA, re_form_iid = NA)

ggplot(p_hybrid, aes(sex, exp(est), fill = prey_group)) +
  facet_wrap(~quarter_f, ncol = 4) +
  geom_bar(stat = "identity")

ggplot(p_hybrid, aes(as.numeric(as.character(quarter_f)), exp(est), fill = prey_group)) +
  facet_wrap(~sex, ncol = 4) +
  geom_bar(stat = "identity")

qqnorm(residuals(h9)); qqline(residuals(h9))

# Quarter effect for aggregated gadid prey by sex at the same reference seal and year.
nd_hybrid_gad <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  sex = unique(diet_model_hybrid$sex),
  prey_group = factor(c(gad_species_keep, "Other Gadidae"), levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_hybrid_gad <- predict(h9, newdata = nd_hybrid_gad, re_form = NA, re_form_iid = NA)

hybrid_gad_quarter <- nd_hybrid_gad |>
  bind_cols(as_tibble(p_hybrid_gad) |> select(est)) |>
  mutate(biomass = exp(est)) |>
  summarise(biomass = sum(biomass), .by = c(quarter_f, sex))

ggplot(hybrid_gad_quarter, aes(as.numeric(as.character(quarter_f)), biomass, color = sex)) +
  geom_line()

plot(ggeffects::ggpredict(h9, c("prey_group", "sex", "quarter_f"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot(ggeffects::ggpredict(h6, c("prey_group", "sex"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot(ggeffects::ggpredict(h6, c("body_length_cm", "prey_group")))
plot(ggeffects::ggpredict(h6, c("body_length_cm", "prey_group", "sex")))

diet_model_hybrid |>
  distinct(sample_id, body_length_cm, sex) |>
  ggplot(aes(body_length_cm, fill = sex)) +
  geom_histogram()

diet_model_hybrid |>
  distinct(sample_id, month, sex) |>
  ggplot(aes(as.factor(month), fill = sex)) +
  geom_bar()
```

## Hybrid biomass predictions per sample

These plots keep predictions on the biomass-per-sample scale instead of converting them to relative composition. This is useful when the question is about absolute prey biomass rather than diet proportion.

```{r hybrid-biomass-predictions}
# Sex effect on biomass/sample from the simpler spatial hybrid model.
nd_bio_sex <- expand_grid(
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_bio_sex <- predict(h6, newdata = nd_bio_sex, re_form = NA, re_form_iid = NA)

bio_sex <- nd_bio_sex |>
  bind_cols(as_tibble(p_bio_sex) |> select(est)) |>
  mutate(biomass = exp(est))

plot_bio_sex_stack <- bio_sex |>
  ggplot(aes(sex, biomass, fill = prey_group)) +
  geom_col(width = 0.6) +
  labs(
    title = "Predicted prey biomass per sample by sex",
    subtitle = paste("Predictions from h6 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = NULL,
    y = "Predicted biomass per sample (kg)",
    fill = "Prey group"
  )

plot_bio_gadus_sex <- bio_sex |>
  filter(prey_group == "Gadus morhua") |>
  ggplot(aes(sex, biomass, fill = sex)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  labs(
    title = "Predicted Gadus morhua biomass per sample by sex",
    subtitle = paste("Predictions from h6 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = NULL,
    y = "Predicted biomass per sample (kg)"
  )

# Length effect on biomass/sample.
length_seq_hybrid <- seq(
  quantile(diet_model_hybrid$body_length_cm, 0.1, na.rm = TRUE),
  quantile(diet_model_hybrid$body_length_cm, 0.9, na.rm = TRUE),
  length.out = 60
)

nd_bio_length <- expand_grid(
  body_length_cm = length_seq_hybrid,
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels)
) |>
  mutate(
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_bio_length <- predict(h6, newdata = nd_bio_length, re_form = NA, re_form_iid = NA)

bio_length <- nd_bio_length |>
  bind_cols(as_tibble(p_bio_length) |> select(est)) |>
  mutate(biomass = exp(est))

plot_bio_length_gadus <- bio_length |>
  filter(prey_group == "Gadus morhua") |>
  ggplot(aes(body_length_cm, biomass, color = sex)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Predicted Gadus morhua biomass per sample over seal length",
    subtitle = paste("Predictions from h6 | Year fixed at", pred_year_hybrid),
    x = "Body length (cm)",
    y = "Predicted biomass per sample (kg)",
    color = "Sex"
  )

# Quarter effect on biomass/sample.
nd_bio_quarter <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_bio_quarter <- predict(h10, newdata = nd_bio_quarter, re_form = NA, re_form_iid = NA)

bio_quarter <- nd_bio_quarter |>
  bind_cols(as_tibble(p_bio_quarter) |> select(est)) |>
  mutate(biomass = exp(est))

plot_bio_quarter_gadus <- bio_quarter |>
  filter(prey_group == "Gadus morhua") |>
  ggplot(aes(as.numeric(as.character(quarter_f)), biomass, color = sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:4) +
  labs(
    title = "Predicted Gadus morhua biomass per sample by quarter",
    subtitle = paste("Predictions from h10 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = "Quarter",
    y = "Predicted biomass per sample (kg)",
    color = "Sex"
  )

plot_bio_quarter_stack <- bio_quarter |>
  ggplot(aes(quarter_f, biomass, fill = prey_group)) +
  geom_col() +
  facet_wrap(~sex) +
  labs(
    title = "Predicted prey biomass per sample by quarter",
    subtitle = paste("Predictions from h10 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = "Quarter",
    y = "Predicted biomass per sample (kg)",
    fill = "Prey group"
  )

# Cod vs whiting over time at a representative location.
nd_cod_whiting_time <- expand_grid(
  year = sort(unique(diet_model_hybrid$year)),
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(c("Gadus morhua", "Merlangius merlangus"), levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_cod_whiting_time <- predict(h6, newdata = nd_cod_whiting_time, re_form = NA, re_form_iid = NA)

cod_whiting_time <- nd_cod_whiting_time |>
  bind_cols(as_tibble(p_cod_whiting_time) |> select(est)) |>
  mutate(biomass = exp(est))

plot_cod_whiting_time <- cod_whiting_time |>
  ggplot(aes(year, biomass, color = prey_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~sex) +
  labs(
    title = "Predicted cod and whiting biomass per sample over time",
    subtitle = paste("Predictions from h6 | Length fixed at", round(pred_length_hybrid, 1), "cm at a representative location"),
    x = "Year",
    y = "Predicted biomass per sample (kg)",
    color = "Prey group"
  )

# Year-space biomass maps from the spatiotemporal model.
map_years_hybrid <- unique(c(min(diet_model_hybrid$year), median(sort(unique(diet_model_hybrid$year))) |> round(), max(diet_model_hybrid$year)))

pg_bio_space <- pred_grid_base |>
  filter(year %in% map_years_hybrid) |>
  expand_grid(
    sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
    prey_group = factor(prey_group_levels, levels = prey_group_levels)
  ) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    year_ct = year - year_center
  )

p_bio_space <- predict(h8, newdata = pg_bio_space)

bio_space <- pg_bio_space |>
  bind_cols(as_tibble(p_bio_space) |> select(est)) |>
  mutate(biomass = exp(est)) |>
  filter(prey_group == "Gadus morhua")

plot_bio_space_gadus <- bio_space |>
  ggplot(aes(X * 1000, Y * 1000, fill = biomass)) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis() +
  coord_sf(
    crs = 32632,
    xlim = range(bio_space$X * 1000),
    ylim = range(bio_space$Y * 1000),
    expand = FALSE
  ) +
  labs(
    title = "Predicted Gadus morhua biomass per sample in space",
    subtitle = paste("Predictions from h8 | Length fixed at", round(pred_length_hybrid, 1), "cm"),
    x = "Longitude",
    y = "Latitude",
    fill = "Biomass\n(kg/sample)"
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(sex ~ year)

plot_bio_sex_stack
plot_bio_gadus_sex
plot_bio_length_gadus
plot_bio_quarter_stack
plot_bio_quarter_gadus
plot_cod_whiting_time
plot_bio_space_gadus
```
