---
title: "Harbour Seal Ontogenetic Diet Hybrid Models"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    page-layout: full
    embed-resources: true
editor: source
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
---

## Purpose

This notebook focuses on ontogenetic variation in diet composition using a hybrid prey-group workflow that splits dominant gadid prey while keeping non-gadid prey grouped at family level.

The workflow is:

- retain colonies through Onsala only,
- retain hunted seals only,
- keep the same dominant non-gadid family grouping as `03b`,
- split `Gadidae` into the most supported scientific-name labels,
- build a hunted-seal sample-by-hybrid-prey-group modelling table,
- test whether seal length changes diet composition,
- then test whether sex adds further structure once length is included.

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(forcats)
library(stringr)
library(janitor)
library(sf)
library(sdmTMB)
library(purrr)
library(rnaturalearth)
library(ggsidekick)

theme_set(theme_sleek())

home <- here::here()

biomass_file <- file.path(home, "Data", "HS_SKAG_estimated_abundance_biomass_130426.csv")

colony_keep <- c("Koster", "Vaderoarna", "Lysekil", "Marstrand", "Onsala")
analysis_year_min <- 2013L
dominant_families <- c("Gadidae", "Clupeidae", "Pleuronectidae", "Gobiidae")

family_palette <- c(
  "Gadidae"        = "#2E6B1F",
  "Clupeidae"      = "#1F5AA6",
  "Pleuronectidae" = "#D94FD3",
  "Gobiidae"       = "#F39C12",
  "Other"          = "#9E4A7A",
  "Total"          = "#000000"
)

normalize_colony_name <- function(x) {
  x <- trimws(as.character(x))
  case_when(
    grepl("der.*arna", x, ignore.case = TRUE) ~ "Vaderoarna",
    grepl("^Hallands.*der", x, ignore.case = TRUE) ~ "Hallands Vadero",
    TRUE ~ x
  )
}
```

## Read and harmonize source data

```{r read-data}
sample_meta_workbook <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year_wb = suppressWarnings(as.integer(year)),
    month_wb = suppressWarnings(as.integer(month)),
    quarter_wb = suppressWarnings(as.integer(quarter)),
    latitude_wb = suppressWarnings(as.numeric(latitude)),
    longitude_wb = suppressWarnings(as.numeric(longitude)),
    colony_wb = normalize_colony_name(colony),
    sex_wb = case_when(
      sex %in% c("Female", "Male") ~ as.character(sex),
      TRUE ~ NA_character_
    ),
    age_wb = suppressWarnings(as.numeric(age)),
    body_length_cm_wb = suppressWarnings(as.numeric(body_length_cm)),
    collection_method_wb = if_else(
      is.na(collection_method) | collection_method == "",
      "Unknown",
      as.character(collection_method)
    )
  )

diet_raw <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  left_join(sample_meta_workbook, by = "sample_id") |>
  mutate(
    year = coalesce(year_wb, suppressWarnings(as.integer(year))),
    month = coalesce(month_wb, suppressWarnings(as.integer(month))),
    quarter = coalesce(quarter_wb, suppressWarnings(as.integer(quarter))),
    latitude = coalesce(latitude_wb, suppressWarnings(as.numeric(latitude))),
    longitude = coalesce(longitude_wb, suppressWarnings(as.numeric(longitude))),
    colony = coalesce(colony_wb, normalize_colony_name(colony)),
    collection_method = coalesce(collection_method_wb, as.character(collection_method)),
    collection_method = if_else(is.na(collection_method) | collection_method == "", "Unknown", collection_method),
    family = if_else(is.na(family) | family == "", "Unknown", family),
    prey_species = prey_species_lat,
    sex = coalesce(sex_wb, case_when(
      sex %in% c("Female", "Male") ~ as.character(sex),
      TRUE ~ NA_character_
    )),
    age = coalesce(age_wb, suppressWarnings(as.numeric(age))),
    body_length_cm = coalesce(body_length_cm_wb, suppressWarnings(as.numeric(body_length_cm))),
    quarter = coalesce(quarter, ((month - 1L) %/% 3L) + 1L)
  ) |>
  select(-ends_with("_wb"))

sample_meta_all <- diet_raw |>
  filter(!is.na(sample_id), !is.na(year), !is.na(latitude), !is.na(longitude)) |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year,
    month,
    quarter,
    lat = latitude,
    lon = longitude,
    collection_method,
    colony,
    sex,
    age,
    body_length_cm
  )

diet_rows_all <- diet_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species),
    prey_species != "",
    tolower(prey_species) != "non prey item"
  ) |>
  transmute(
    sample_id,
    family,
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, family))

sample_totals_all <- diet_rows_all |>
  summarise(sum_biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = sample_id)

sample_meta_filtered <- sample_meta_all |>
  left_join(sample_totals_all, by = "sample_id") |>
  mutate(sum_biomass_kg = coalesce(sum_biomass_kg, 0)) |>
  filter(
    collection_method == "Hunt",
    colony %in% colony_keep,
    sum_biomass_kg > 0
  )

if (nrow(sample_meta_filtered) == 0) {
  stop("No hunted samples remain after filtering by colony and positive total biomass.")
}

analysis_year_max <- max(sample_meta_filtered$year, na.rm = TRUE)

sample_meta_main <- sample_meta_filtered |>
  filter(year >= analysis_year_min, year <= analysis_year_max)

diet_rows_main <- diet_rows_all |>
  semi_join(sample_meta_main, by = "sample_id")
```

## Support summaries and family grouping

```{r support-tables}
n_main_samples <- nrow(sample_meta_main)

family_levels <- unique(c(dominant_families, "Other"))

family_support_main <- diet_rows_main |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = family
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence
  ) |>
  arrange(desc(total_biomass_kg))

diet_rows_grouped <- diet_rows_main |>
  mutate(family = if_else(family %in% dominant_families, family, "Other")) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, family))

family_support_grouped <- diet_rows_grouped |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = family
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence,
    family = factor(family, levels = family_levels)
  ) |>
  arrange(family)
```

## Build hunted-only model table

```{r build-model-table}
year_center <- mean(sample_meta_main$year, na.rm = TRUE)

diet_model_all <- sample_meta_main |>
  select(sample_id, year, month, quarter, lat, lon, colony, sum_biomass_kg, sex, age, body_length_cm) |>
  expand_grid(family = family_levels) |>
  left_join(diet_rows_grouped, by = c("sample_id", "family")) |>
  mutate(
    biomass_kg = coalesce(biomass_kg, 0),
    present = as.integer(biomass_kg > 0),
    prop_biomass = if_else(sum_biomass_kg > 0, biomass_kg / sum_biomass_kg, 0),
    year_ct = year - year_center,
    colony = factor(colony, levels = colony_keep),
    family = factor(family, levels = family_levels),
    sex = factor(sex, levels = c("Female", "Male")),
    year_f = factor(year),
    quarter_f = factor(quarter),
    month_f = factor(month)
  ) |>
  add_utm_columns(ll_names = c("lon", "lat"))

diet_model <- diet_model_all |>
  filter(!is.na(body_length_cm))

diet_model_sex <- diet_model |>
  filter(!is.na(sex))

diet_model |>
  summarise(
    n_samples = n_distinct(sample_id),
    n_rows = n(),
    min_year = min(year),
    max_year = max(year),
    n_families = n_distinct(family),
    length_min = min(body_length_cm),
    length_max = max(body_length_cm)
  )
```

## Support diagnostics

```{r support-plots}
hunt_samples_by_year <- sample_meta_main |>
  count(year, name = "n_samples")

missing_covariates <- sample_meta_main |>
  summarise(
    n_samples = n(),
    n_length = sum(!is.na(body_length_cm)),
    n_sex = sum(!is.na(sex)),
    n_age = sum(!is.na(age)),
    n_length_sex = sum(!is.na(body_length_cm) & !is.na(sex)),
    n_all_three = sum(!is.na(body_length_cm) & !is.na(sex) & !is.na(age))
  )

age_length_plot <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(age), !is.na(sex)) |>
  ggplot(aes(age, body_length_cm, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    title = "Relation between seal age and length",
    subtitle = "Separate fitted lines for female and male hunted seals",
    x = "Age",
    y = "Body length (cm)",
    color = "Sex"
  )

length_histogram <- sample_meta_main |>
  filter(!is.na(body_length_cm)) |>
  ggplot(aes(body_length_cm)) +
  geom_histogram(binwidth = 5, color = "white", fill = "grey40") +
  labs(
    title = "Seal length distribution in hunted samples",
    x = "Body length (cm)",
    y = "Samples"
  )

length_histogram_sex_colony <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(sex)) |>
  mutate(colony = factor(colony, levels = colony_keep)) |>
  ggplot(aes(body_length_cm, fill = sex)) +
  geom_histogram(binwidth = 5, color = "white") +
  facet_wrap(~colony, ncol = 1, scales = "free_y") +
  labs(
    title = "Seal length distribution in hunted samples by colony and sex",
    subtitle = "Colonies ordered from north to south",
    x = "Body length (cm)",
    y = "Samples",
    fill = "Sex"
  )

length_by_year <- sample_meta_main |>
  filter(!is.na(body_length_cm)) |>
  ggplot(aes(year, body_length_cm)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Seal length over time",
    x = "Year",
    y = "Body length (cm)"
  )

sex_balance <- sample_meta_main |>
  filter(!is.na(sex)) |>
  count(sex) |>
  ggplot(aes(sex, n, fill = sex)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Sex balance in hunted samples",
    x = NULL,
    y = "Samples"
  )

sex_balance_colony <- sample_meta_main |>
  filter(!is.na(sex)) |>
  mutate(colony = factor(colony, levels = colony_keep)) |>
  count(colony, sex, name = "n") |>
  group_by(colony) |>
  mutate(
    total_n = sum(n),
    prop = n / total_n
  ) |>
  ungroup() |>
  ggplot(aes(colony, prop, fill = sex)) +
  geom_col() +
  geom_text(
    data = ~ distinct(.x, colony, total_n),
    aes(x = colony, y = 1.02, label = total_n),
    inherit.aes = FALSE,
    vjust = 0
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1.08),
    expand = c(0, 0)
  ) +
  labs(
    title = "Relative sex composition by colony",
    subtitle = "Numbers above bars show total hunted samples with known sex",
    x = NULL,
    y = "Relative abundance",
    fill = "Sex"
  )

family_support_plot <- family_support_grouped |>
  mutate(family = fct_reorder(as.character(family), total_biomass_kg)) |>
  ggplot(aes(total_biomass_kg, family)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Grouped prey families retained for modelling",
    subtitle = "Same dominant-family set as 03b",
    x = "Total biomass (kg)",
    y = NULL
  )

missing_covariates
hunt_samples_by_year
age_length_plot
length_histogram
length_histogram_sex_colony
length_by_year
sex_balance
sex_balance_colony
family_support_plot
```

## Prediction grid

```{r prediction-grid}
d <- diet_model

hull_idx <- chull(d$X, d$Y)
hull_coords <- cbind(d$X[hull_idx], d$Y[hull_idx])
hull_coords <- rbind(hull_coords, hull_coords[1, ])

sp_poly <- sp::SpatialPolygons(
  list(sp::Polygons(list(sp::Polygon(hull_coords)), ID = 1))
)

sp_poly_df <- sp::SpatialPolygonsDataFrame(
  sp_poly,
  data = data.frame(ID = 1)
)

cell_width <- 0.5

pred_grid_base <- expand.grid(
  X = seq(min(d$X), max(d$X), cell_width),
  Y = seq(min(d$Y), max(d$Y), cell_width),
  year = seq(min(d$year), max(d$year))
)

sp::coordinates(pred_grid_base) <- c("X", "Y")

inside <- !is.na(sp::over(pred_grid_base, as(sp_poly_df, "SpatialPolygons")))

pred_grid_base <- pred_grid_base[inside, ]

pred_grid_base <- as.data.frame(pred_grid_base) |>
  mutate(
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model$year_f))
  )

land <- ne_countries(scale = "large", returnclass = "sf") |>
  st_transform(32632)
```

## Mesh

```{r mesh}
mesh_cutoff_km <- 0.5

mesh <- make_mesh(diet_model, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

plot(mesh)
```

## Shared prediction reference

```{r shared-prediction-reference}
prediction_reference <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(sex)) |>
  summarise(
    pred_year = round(median(sort(unique(year)))),
    pred_length_cm = median(body_length_cm, na.rm = TRUE)
  )

pred_year_common <- prediction_reference$pred_year
pred_length_common <- prediction_reference$pred_length_cm
```

## Hybrid prey-group questions: sex, length, year, space

This section uses one hybrid prey grouping for the hunted samples:

- split `Gadidae` into the most supported scientific-name labels,
- keep the remaining gadids as `Other Gadidae`,
- keep non-gadid prey at the same grouped family level used above.

```{r hybrid-prey-data}
gad_species_keep <- c(
  "Gadus morhua",
  "Merlangius merlangus",
  "Pollachius sp. / Melanogrammus aeglefinus",
  "Gadidae"
)

prey_group_levels <- c(
  gad_species_keep,
  "Other Gadidae",
  "Clupeidae",
  "Pleuronectidae",
  "Gobiidae",
  "Other"
)

hybrid_rows <- diet_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species),
    prey_species != "",
    tolower(prey_species) != "non prey item"
  ) |>
  transmute(
    sample_id,
    prey_group = case_when(
      family == "Gadidae" & prey_species %in% gad_species_keep ~ prey_species,
      family == "Gadidae" ~ "Other Gadidae",
      family == "Clupeidae" ~ "Clupeidae",
      family == "Pleuronectidae" ~ "Pleuronectidae",
      family == "Gobiidae" ~ "Gobiidae",
      TRUE ~ "Other"
    ),
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, prey_group))

prey_group_support <- hybrid_rows |>
  semi_join(sample_meta_main |> filter(!is.na(body_length_cm), !is.na(sex)), by = "sample_id") |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = prey_group
  ) |>
  mutate(prey_group = factor(prey_group, levels = prey_group_levels)) |>
  arrange(prey_group)

diet_model_hybrid <- sample_meta_main |>
  filter(!is.na(body_length_cm), !is.na(sex)) |>
  select(sample_id, year, month, quarter, lat, lon, colony, sum_biomass_kg, sex, body_length_cm) |>
  expand_grid(prey_group = prey_group_levels) |>
  left_join(hybrid_rows, by = c("sample_id", "prey_group")) |>
  mutate(
    biomass_kg = coalesce(biomass_kg, 0),
    year_ct = year - year_center,
    year_f = factor(year),
    quarter_f = factor(quarter),
    prey_group = factor(prey_group, levels = prey_group_levels),
    sex = factor(sex, levels = c("Female", "Male"))
  ) |>
  add_utm_columns(ll_names = c("lon", "lat"))

prey_group_support
```

```{r fit-hybrid-models}
mesh_hybrid <- make_mesh(diet_model_hybrid, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

diet_model_hybrid$body_length_cm_sc <- (diet_model_hybrid$body_length_cm - mean(diet_model_hybrid$body_length_cm)) / sd(diet_model_hybrid$body_length_cm)

h1 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm_sc + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h2 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm_sc * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h3 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + s(body_length_cm_sc, by = prey_group) + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h4 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm_sc * sex + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "on"
)

h5 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm_sc + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h6 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm_sc * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h7 <- sdmTMB(
  formula = biomass_kg ~ body_length_cm * prey_group * sex + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h8 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + body_length_cm_sc * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatiotemporal = "iid",
  time = "year",
  spatial_varying = ~ 0 + prey_group
)

h9 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + quarter_f * sex + body_length_cm_sc * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h10 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + quarter_f + 
    body_length_cm * prey_group + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

h11 <- sdmTMB(
  formula = biomass_kg ~ sex * prey_group + quarter_f * sex + body_length_cm_sc * prey_group + body_length_cm_sc * sex + (1 | year_f),
  data = diet_model_hybrid,
  mesh = mesh_hybrid,
  family = tweedie(),
  spatial = "off",
  spatial_varying = ~ 0 + prey_group
)

AIC(h9, h11)
AIC(h1, h2, h3, h4, h5, h6, h7, h8, h9, h10) |>
  arrange(AIC)

# Maybe prey*quarter as well? Then we might just do a 3-way interaction, which may or may not be supported..
```

```{r}
library(tidyverse)
library(broom.mixed)  # or broom if h9 is not mixed

# biomass_kg ~ sex * prey_group + quarter_f * sex + body_length_cm * prey_group + (1 | year_f)
# tidy(h9) |> as.data.frame()
# 
# # female, q1?
# ggeffects::ggpredict(h9, terms = c("quarter_f", "sex"))
# 
# ggeffects::ggpredict(h9, terms = c("quarter_f", "sex")) |>
#   plot()

```

```{r hybrid-level-diagnostics}
pred_year_hybrid <- pred_year_common
pred_length_hybrid <- pred_length_common

# Season effects from the hybrid model using the same reference seal and year as the family plots.
nd_hybrid <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels),
  sex = unique(diet_model_hybrid$sex),
  body_length_cm = c(80, 110, 140)
) |>
  mutate(
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_hybrid <- predict(h11, newdata = nd_hybrid, re_form = NA, re_form_iid = NA)

fig_sex_quarter <- ggplot(p_hybrid, aes(sex, exp(est), fill = prey_group)) +
  facet_grid(body_length_cm ~ quarter_f) +
  geom_bar(stat = "identity")
fig_sex_quarter

ggplot(p_hybrid, aes(as.numeric(as.character(quarter_f)), exp(est), fill = prey_group)) +
  facet_wrap(~sex, ncol = 4) +
  geom_bar(stat = "identity")

fig_qqplot <- ggplot(data.frame(res = residuals(h11)), aes(sample = res)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantiles", y = "Sample quantiles",
       title = "Q-Q plot of model residuals")
fig_qqplot

# Quarter effect for aggregated gadid prey by sex at the same reference seal and year.
nd_hybrid_gad <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  sex = unique(diet_model_hybrid$sex),
  prey_group = factor(c(gad_species_keep, "Other Gadidae"), levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_hybrid_gad <- predict(h11, newdata = nd_hybrid_gad, re_form = NA, re_form_iid = NA)

hybrid_gad_quarter <- nd_hybrid_gad |>
  bind_cols(as_tibble(p_hybrid_gad) |> select(est)) |>
  mutate(biomass = exp(est)) |>
  summarise(biomass = sum(biomass), .by = c(quarter_f, sex))

ggplot(hybrid_gad_quarter, aes(as.numeric(as.character(quarter_f)), biomass, color = sex)) +
  geom_line()

plot(ggeffects::ggpredict(h11, c("sex", "prey_group", "quarter_f"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggeffects::ggpredict(h11, c("sex"))
ggeffects::ggpredict(h11, c("sex", "prey_group"))
plot(ggeffects::ggpredict(h11, c("sex", "prey_group"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

plot(ggeffects::ggpredict(h9, c("body_length_cm_sc", "prey_group")))

plot(ggeffects::ggpredict(h9, c("body_length_cm_sc", "prey_group", "sex")))

diet_model_hybrid |>
  distinct(sample_id, body_length_cm, sex) |>
  ggplot(aes(body_length_cm, fill = sex)) +
  geom_histogram()

diet_model_hybrid |>
  distinct(sample_id, month, sex) |>
  ggplot(aes(as.factor(month), fill = sex)) +
  geom_bar()
```

## Predict with sims to get uncertainty on prey, sex and quarter

```{r}
# formula = biomass_kg ~ sex * prey_group + quarter_f * sex + body_length_cm_sc * prey_group + body_length_cm_sc * sex + (1 | year_f)

nd_sex <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels),
  body_length_cm_sc = seq(
    min(diet_model_hybrid$body_length_cm_sc), 
    max(diet_model_hybrid$body_length_cm_sc),
    length.out = 30),
  sex = unique(diet_model_hybrid$sex)
)

# predict without sims
pred <- predict(
  h9, newdata = nd_sex, 
  re_form = NA, re_form_iid = NA
  )

pred_sum <- pred |> 
  summarise(sum_exp_est = sum(exp(est)), .by = c(quarter_f, body_length_cm_sc, sex))

ggplot(pred_sum, aes(body_length_cm_sc, sum_exp_est, color = sex)) + 
  geom_line() + 
  facet_wrap(~quarter_f)

# Now don't summarise and instead plot by prey group
# Predict without sims. Do it for a single size?
nd_sex_nosize <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels),
  sex = unique(diet_model_hybrid$sex)
) |> 
  mutate(body_length_cm_sc = 0)

pred <- predict(
  h9, newdata = nd_sex_nosize, 
  se_fit = TRUE,
  re_form = NA, re_form_iid = NA
  )

ggplot(pred, aes(exp(est), reorder(prey_group, exp(est)), color = quarter_f)) + 
  geom_point(position = position_dodge(width = 0.4)) + 
  geom_errorbar(aes(xmin = exp(est - est_se*1.96),
                    xmax = exp(est + est_se*1.96)),
                width = 0, position = position_dodge(width = 0.4)) +
  facet_wrap(~sex) +
  labs(x = "Prey weight [g]", color = "Quarter") +
  scale_color_brewer(palette = "Dark2") +
  guides(color = guide_legend(position = "inside")) +
  theme(axis.title.y = element_blank(),
        legend.position.inside = c(0.95, 0.15)) 
  
# Try with model h9 - sex*size interaction is not massive.
# pred_h9 <- predict(
#   h9, newdata = nd_sex, 
#   re_form = NA, re_form_iid = NA
#   )
# 
# pred_sum_h9 <- pred_h9 |> 
#   summarise(sum_exp_est = sum(exp(est)), .by = c(quarter_f, body_length_cm_sc, sex))
# 
# ggplot(pred_sum_h9, aes(body_length_cm_sc, sum_exp_est, color = sex)) + 
#   geom_line() + 
#   facet_wrap(~quarter_f)

# Predict the same with sims
# nd_hybrid_long <- expand_grid(
#   quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
#   prey_group = factor(prey_group_levels, levels = prey_group_levels),
#   body_length_cm = seq(
#     min(diet_model_hybrid$body_length_cm_sc), 
#     max(diet_model_hybrid$body_length_cm_sc),
#     length.out = 30),
#   sex = unique(diet_model_hybrid$sex)
# )

pred_sims <- predict(
  h9, newdata = nd_sex, 
  re_form = NA, re_form_iid = NA, 
  nsim = 1000) |> 
  as.data.frame(row.names = FALSE)

pred_sims_df <- nd_sex |> 
  bind_cols(pred_sims)

plot_sim <- pred_sims_df |> 
  pivot_longer("V1":"V1000", names_to = "sim", values_to = "est") |> 
  summarise(sum_exp_est = sum(exp(est)), .by = c(quarter_f, body_length_cm_sc, sex, sim)) |> 
  summarise(
    lo = quantile(sum_exp_est, 0.025),
    hi = quantile(sum_exp_est, 0.975),
    sum_exp_est = mean(sum_exp_est),
    .by = c(quarter_f, body_length_cm_sc, sex))
  # pivot_wider(names_from = sex, values_from = est) |>
  # mutate(diff = Female - Male)

fig_length_composition <- ggplot(plot_sim, aes(body_length_cm_sc, sum_exp_est, color = sex, fill = sex)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.3, color = NA) +
  facet_wrap(~quarter_f)
fig_length_composition

# Now calculate the differences between male and female
plot_sim <- pred_sims_df |> 
  pivot_longer("V1":"V1000", names_to = "sim", values_to = "est") |> 
  summarise(sum_exp_est = sum(exp(est)), .by = c(quarter_f, body_length_cm_sc, sex, sim)) |> 
  pivot_wider(names_from = sex, values_from = sum_exp_est) |>
  mutate(diff = Female - Male) |> 
  summarise(
    mean_diff = mean(diff),
    lo = quantile(diff, 0.025),
    hi = quantile(diff, 0.975),
    .by = c(body_length_cm_sc, quarter_f)
    )

ggplot(plot_sim, aes(body_length_cm_sc, mean_diff)) + 
  geom_hline(yintercept = 0, lty = 2) +
  geom_line() + 
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.3, color = NA) +
  facet_wrap(~quarter_f)

# Significance between male and female meaningful?
diet_model_hybrid$body_length_cm_sc <- (diet_model_hybrid$body_length_cm - mean(diet_model_hybrid$body_length_cm)) / sd(diet_model_hybrid$body_length_cm)

plot_sim <- pred_sims_df |> 
  pivot_longer("V1":"V1000", names_to = "sim", values_to = "est") |> 
  mutate(body_length_cm = body_length_cm_sc * sd(diet_model_hybrid$body_length_cm) + mean(diet_model_hybrid$body_length_cm)) |> 
  mutate(size_cat = ifelse(body_length_cm > 120, "Adult", "Juvenile")) |> 
  summarise(sum_exp_est = sum(exp(est)), .by = c(quarter_f, sex, size_cat, sim)) |> 
  pivot_wider(names_from = sex, values_from = sum_exp_est) |>
  mutate(diff = Female - Male)

labs <- plot_sim |>
  summarise(
    p_fem_gt = mean(diff > 0),
    .by = c(quarter_f, size_cat)
  )

fig_simulations <- ggplot(plot_sim, aes(as.numeric(as.character(quarter_f)), diff)) +
  geom_hline(yintercept = 0, lty = 2) +
  facet_wrap(~size_cat) +
  geom_jitter(data = plot_sim |>
                group_by(quarter_f, size_cat) |>
                sample_n(100),
              aes(as.numeric(as.character(quarter_f)), diff),
              alpha = 0.2, width = 0.2, height = 0) +
  geom_boxplot(aes(group = quarter_f), fill = NA, outlier.color = NA, width = 0.2) +
  geom_violin(aes(group = quarter_f), fill = NA) +
  geom_text(
    data = labs,
    aes(x = as.numeric(as.character(quarter_f)), y = Inf,
        label = paste0("p = ", round(p_fem_gt, 2))),
    vjust = 1.5, inherit.aes = FALSE
  ) +
  labs(x = "Quarter", y = "Difference in biomass (Female - Male)")
fig_simulations
```

## Hybrid biomass predictions per sample

These plots keep predictions on the biomass-per-sample scale instead of converting them to relative composition. This is useful when the question is about absolute prey biomass rather than diet proportion.

```{r hybrid-biomass-predictions}
# Sex effect on biomass/sample from the simpler spatial hybrid model.
nd_bio_sex <- expand_grid(
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_bio_sex <- predict(h9, newdata = nd_bio_sex, re_form = NA, re_form_iid = NA)

bio_sex <- nd_bio_sex |>
  bind_cols(as_tibble(p_bio_sex) |> select(est)) |>
  mutate(biomass = exp(est))

plot_bio_sex_stack <- bio_sex |>
  ggplot(aes(sex, biomass, fill = prey_group)) +
  geom_col(width = 0.6) +
  labs(
    title = "Predicted prey biomass per sample by sex",
    subtitle = paste("Predictions from h6 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = NULL,
    y = "Predicted biomass per sample (kg)",
    fill = "Prey group"
  )

plot_bio_gadus_sex <- bio_sex |>
  filter(prey_group == "Gadus morhua") |>
  ggplot(aes(sex, biomass, fill = sex)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  labs(
    title = "Predicted Gadus morhua biomass per sample by sex",
    subtitle = paste("Predictions from h6 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = NULL,
    y = "Predicted biomass per sample (kg)"
  )

# Length effect on biomass/sample.
length_seq_hybrid <- seq(
  quantile(diet_model_hybrid$body_length_cm, 0.1, na.rm = TRUE),
  quantile(diet_model_hybrid$body_length_cm, 0.9, na.rm = TRUE),
  length.out = 60
)

nd_bio_length <- expand_grid(
  body_length_cm = length_seq_hybrid,
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels)
) |>
  mutate(
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_bio_length <- predict(h9, newdata = nd_bio_length, re_form = NA, re_form_iid = NA)

bio_length <- nd_bio_length |>
  bind_cols(as_tibble(p_bio_length) |> select(est)) |>
  mutate(biomass = exp(est))

plot_bio_length_gadus <- bio_length |>
  filter(prey_group == "Gadus morhua") |>
  ggplot(aes(body_length_cm, biomass, color = sex)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Predicted Gadus morhua biomass per sample over seal length",
    subtitle = paste("Predictions from h6 | Year fixed at", pred_year_hybrid),
    x = "Body length (cm)",
    y = "Predicted biomass per sample (kg)",
    color = "Sex"
  )

# Quarter effect on biomass/sample.
nd_bio_quarter <- expand_grid(
  quarter_f = factor(c(1:4), levels = levels(diet_model_hybrid$quarter_f)),
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(prey_group_levels, levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year = pred_year_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_bio_quarter <- predict(h9, newdata = nd_bio_quarter, re_form = NA, re_form_iid = NA)

bio_quarter <- nd_bio_quarter |>
  bind_cols(as_tibble(p_bio_quarter) |> select(est)) |>
  mutate(biomass = exp(est))

plot_bio_quarter_gadus <- bio_quarter |>
  filter(prey_group == "Gadus morhua") |>
  ggplot(aes(as.numeric(as.character(quarter_f)), biomass, color = sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:4) +
  labs(
    title = "Predicted Gadus morhua biomass per sample by quarter",
    subtitle = paste("Predictions from h10 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = "Quarter",
    y = "Predicted biomass per sample (kg)",
    color = "Sex"
  )

plot_bio_quarter_stack <- bio_quarter |>
  ggplot(aes(quarter_f, biomass, fill = prey_group)) +
  geom_col() +
  facet_wrap(~sex) +
  labs(
    title = "Predicted prey biomass per sample by quarter",
    subtitle = paste("Predictions from h10 | Length fixed at", round(pred_length_hybrid, 1), "cm and year", pred_year_hybrid),
    x = "Quarter",
    y = "Predicted biomass per sample (kg)",
    fill = "Prey group"
  )

# Cod vs whiting over time at a representative location.
nd_cod_whiting_time <- expand_grid(
  year = sort(unique(diet_model_hybrid$year)),
  sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
  prey_group = factor(c("Gadus morhua", "Merlangius merlangus"), levels = prey_group_levels)
) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    year_ct = year - year_center,
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    X = median(diet_model_hybrid$X, na.rm = TRUE),
    Y = median(diet_model_hybrid$Y, na.rm = TRUE)
  )

p_cod_whiting_time <- predict(h9, newdata = nd_cod_whiting_time, re_form = NA, re_form_iid = NA)

cod_whiting_time <- nd_cod_whiting_time |>
  bind_cols(as_tibble(p_cod_whiting_time) |> select(est)) |>
  mutate(biomass = exp(est))

plot_cod_whiting_time <- cod_whiting_time |>
  ggplot(aes(year, biomass, color = prey_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~sex) +
  labs(
    title = "Predicted cod and whiting biomass per sample over time",
    subtitle = paste("Predictions from h6 | Length fixed at", round(pred_length_hybrid, 1), "cm at a representative location"),
    x = "Year",
    y = "Predicted biomass per sample (kg)",
    color = "Prey group"
  )

# Year-space biomass maps from the spatiotemporal model.
map_years_hybrid <- unique(c(min(diet_model_hybrid$year), median(sort(unique(diet_model_hybrid$year))) |> round(), max(diet_model_hybrid$year)))

pg_bio_space <- pred_grid_base |>
  filter(year %in% map_years_hybrid) |>
  expand_grid(
    sex = factor(c("Female", "Male"), levels = levels(diet_model_hybrid$sex)),
    prey_group = factor(prey_group_levels, levels = prey_group_levels)
  ) |>
  mutate(
    body_length_cm = pred_length_hybrid,
    quarter_f = factor(1, levels = levels(diet_model_hybrid$quarter_f)),
    year_f = factor(year, levels = levels(diet_model_hybrid$year_f)),
    year_ct = year - year_center
  )

p_bio_space <- predict(h9, newdata = pg_bio_space)

bio_space <- pg_bio_space |>
  bind_cols(as_tibble(p_bio_space) |> select(est)) |>
  mutate(biomass = exp(est)) |>
  filter(prey_group == "Gadus morhua")

plot_bio_space_gadus <- bio_space |>
  ggplot(aes(X * 1000, Y * 1000, fill = biomass)) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "fourth_root_power") +
  coord_sf(
    crs = 32632,
    xlim = range(bio_space$X * 1000),
    ylim = range(bio_space$Y * 1000),
    expand = FALSE
  ) +
  labs(
    title = "Predicted Gadus morhua biomass per sample in space",
    subtitle = paste("Predictions from h8 | Length fixed at", round(pred_length_hybrid, 1), "cm"),
    x = "Longitude",
    y = "Latitude",
    fill = "Biomass\n(kg/sample)"
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_grid(sex ~ year)

# SVC field
plot_svc <- p_bio_space |>
  filter(year == min(year)) |> 
  pivot_longer(starts_with("zeta"), names_to = "prey", values_to = "zeta") |> 
  mutate(prey = str_remove(prey, "zeta_s_prey_group")) |> 
  ggplot(aes(X * 1000, Y * 1000, fill = zeta)) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  #viridis::scale_fill_viridis(trans = "fourth_root_power") +
  scale_fill_gradient2() + 
  coord_sf(
    crs = 32632,
    xlim = range(bio_space$X * 1000),
    ylim = range(bio_space$Y * 1000),
    expand = FALSE
  ) +
  labs(
    title = "Predicted Gadus morhua biomass per sample in space",
    #subtitle = paste("Predictions from h8 | Length fixed at", round(pred_length_hybrid, 1), "cm"),
    x = "Longitude",
    y = "Latitude",
    fill = "SVC"
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ prey, ncol = 5)

plot_bio_sex_stack
plot_bio_gadus_sex
plot_bio_length_gadus
plot_bio_quarter_stack
plot_bio_quarter_gadus
plot_cod_whiting_time
plot_bio_space_gadus
plot_svc
```

## Save paper figures

```{r save-figures}
fig_dir <- here::here("paper", "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

# Figure 1 – stacked prey biomass bars by sex, body length x quarter
ggsave(file.path(fig_dir, "sex_quarter.pdf"),
       fig_sex_quarter, width = 9, height = 7)

# Figure 2 – total biomass ribbon over scaled body length by quarter
ggsave(file.path(fig_dir, "length_composition.pdf"),
       fig_length_composition, width = 9, height = 6)

# Figure 3 – Female minus Male difference boxplot by quarter x size class
ggsave(file.path(fig_dir, "simulations.pdf"),
       fig_simulations, width = 8, height = 5)

# Figure 4 – Gadus morhua biomass by quarter and sex (named object)
ggsave(file.path(fig_dir, "gadus_quarter.pdf"),
       plot_bio_quarter_gadus, width = 7, height = 5)

# Figure 5 – QQ plot of model residuals (ggplot2 version)
pdf(file.path(fig_dir, "qqplot.pdf"), width = 5, height = 5)
print(fig_qqplot)
dev.off()
```
