---
title: "Harbour Seal Diet Family Models"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    page-layout: full
    embed-resources: true
editor: source
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
---

## Purpose

This notebook builds a dedicated prey-family modelling pipeline for the harbour seal diet data.

The workflow is:

-   retain colonies through Onsala only,
-   retain Faeces and Hunt samples only,
-   build a sample-by-family modelling table with zero expansion,
-   use 2013-2025 as the main analysis period, previous period too few samples restricted in time space and just scats. The early years remain available in the filtered metadata and can be used later for a sensitivity appendix if needed.
-   model prey biomass by family with collection_method and family-specific smooth year effects

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(ggplot2)
library(forcats)
library(stringr)
library(janitor)
library(sf)
library(sdmTMB)
library(purrr)
library(rnaturalearth)
library(sf)
library(ggsidekick)

theme_set(theme_sleek())

home <- here::here()

biomass_file <- file.path(home, "Data", "HS_SKAG_estimated_abundance_biomass_130426.csv")
ices_areas_file <- file.path(home, "Data", "shapefiles", "ICES_areas", "ICES_Areas_20160601_cut_dense_3857.shp")

colony_keep <- c("Koster", "Väderöarna", "Lysekil", "Marstrand", "Onsala")
method_keep <- c("Faeces", "Hunt")
analysis_year_min <- 2013L
dominant_families <- c("Gadidae", "Clupeidae", "Pleuronectidae", "Gobiidae")

family_palette <- c(
  "Gadidae"        = "#2E6B1F",
  "Clupeidae"      = "#1F5AA6",
  "Pleuronectidae" = "#D94FD3",
  "Gobiidae"       = "#F39C12",
  "Other"          = "#9E4A7A",
  "Total"          = "#000000"
)

normalize_colony_name <- function(x) {
  x <- trimws(as.character(x))
  case_when(
    grepl("der.*arna", x, ignore.case = TRUE) ~ "Väderöarna",
    grepl("^Hallands.*der", x, ignore.case = TRUE) ~ "Hallands Väderö",
    TRUE ~ x
  )
}
```

## Read and harmonize source data

```{r read-data}
sample_meta_workbook <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year_wb = suppressWarnings(as.integer(year)),
    month_wb = suppressWarnings(as.integer(month)),
    quarter_wb = suppressWarnings(as.integer(quarter)),
    latitude_wb = suppressWarnings(as.numeric(latitude)),
    longitude_wb = suppressWarnings(as.numeric(longitude)),
    colony_wb = normalize_colony_name(colony),
    collection_method_wb = if_else(
      is.na(collection_method) | collection_method == "",
      "Unknown",
      as.character(collection_method)
    )
  )

diet_raw <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names() |>
  left_join(sample_meta_workbook, by = "sample_id") |>
  mutate(
    year = coalesce(year_wb, suppressWarnings(as.integer(year))),
    month = coalesce(month_wb, suppressWarnings(as.integer(month))),
    quarter = coalesce(quarter_wb, suppressWarnings(as.integer(quarter))),
    latitude = coalesce(latitude_wb, suppressWarnings(as.numeric(latitude))),
    longitude = coalesce(longitude_wb, suppressWarnings(as.numeric(longitude))),
    colony = coalesce(colony_wb, normalize_colony_name(colony)),
    collection_method = coalesce(collection_method_wb, as.character(collection_method)),
    collection_method = if_else(is.na(collection_method) | collection_method == "", "Unknown", collection_method),
    colony = if_else(is.na(colony) | colony == "", "Unknown", colony),
    family = if_else(is.na(family) | family == "", "Unknown", family),
    prey_species = prey_species_lat,
    quarter = coalesce(quarter, ((month - 1L) %/% 3L) + 1L)
  ) |>
  select(-ends_with("_wb"))

sample_meta_all <- diet_raw |>
  filter(!is.na(sample_id), !is.na(year), !is.na(latitude), !is.na(longitude)) |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year,
    month,
    quarter,
    lat = latitude,
    lon = longitude,
    collection_method,
    colony
  )

diet_rows_all <- diet_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species),
    prey_species != "",
    tolower(prey_species) != "non prey item"
  ) |>
  transmute(
    sample_id,
    family,
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, family))

sample_totals_all <- diet_rows_all |>
  summarise(sum_biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = sample_id)

sample_meta_filtered <- sample_meta_all |>
  left_join(sample_totals_all, by = "sample_id") |>
  mutate(sum_biomass_kg = coalesce(sum_biomass_kg, 0)) |>
  filter(
    collection_method %in% method_keep,
    colony %in% colony_keep,
    sum_biomass_kg > 0
  )

if (nrow(sample_meta_filtered) == 0) {
  stop("No samples remain after filtering by colony, method, and positive total biomass.")
}

analysis_year_max <- max(sample_meta_filtered$year, na.rm = TRUE)

sample_meta_main <- sample_meta_filtered |>
  filter(year >= analysis_year_min, year <= analysis_year_max)

diet_rows_main <- diet_rows_all |>
  semi_join(sample_meta_main, by = "sample_id")
```

## Support summaries and family grouping

```{r support-tables}
n_main_samples <- nrow(sample_meta_main)

family_support_main <- diet_rows_main |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = family
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence
  ) |>
  arrange(desc(total_biomass_kg))

family_keep <- dominant_families
family_levels <- unique(c(family_keep, "Other"))

diet_rows_grouped <- diet_rows_main |>
  mutate(family = if_else(family %in% family_keep, family, "Other")) |>
  summarise(biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = c(sample_id, family))

family_support_grouped <- diet_rows_grouped |>
  summarise(
    total_biomass_kg = sum(biomass_kg, na.rm = TRUE),
    n_samples_present = n_distinct(sample_id[biomass_kg > 0]),
    .by = family
  ) |>
  mutate(
    prevalence = n_samples_present / n_main_samples,
    zero_rate = 1 - prevalence,
    family = factor(family, levels = family_levels)
  ) |>
  arrange(family)

#write_csv(sample_meta_filtered, file.path(data_dir, "sample-metadata-filtered-all-years.csv"))
#write_csv(sample_meta_main, file.path(data_dir, "sample-metadata-main-period.csv"))
#write_csv(family_support_main, file.path(tables_dir, "family-support-main-period-raw.csv"))
#write_csv(family_support_grouped, file.path(tables_dir, "family-support-main-period-grouped.csv"))
```

```{r build-model-table}
year_center <- mean(sample_meta_main$year, na.rm = TRUE)

diet_model <- sample_meta_main |>
  select(sample_id, year, month, quarter, lat, lon, collection_method, colony, sum_biomass_kg) |>
  tidyr::expand_grid(family = family_levels) |>
  left_join(diet_rows_grouped, by = c("sample_id", "family")) |>
  mutate(
    biomass_kg = coalesce(biomass_kg, 0),
    present = as.integer(biomass_kg > 0),
    prop_biomass = if_else(sum_biomass_kg > 0, biomass_kg / sum_biomass_kg, 0),
    year_ct = year - year_center,
    collection_method = factor(collection_method, levels = method_keep),
    colony = factor(colony, levels = colony_keep),
    family = factor(family, levels = family_levels),
    year_f = as.factor(year),
    quarter_f = factor(quarter),
    month_f = factor(month)
  ) |>
  add_utm_columns(ll_names = c("lon", "lat"))

#write_csv(diet_model, file.path(data_dir, "diet-family-model-table-2013-2025.csv"))

diet_model |>
  summarise(
    n_samples = n_distinct(sample_id),
    n_rows = n(),
    min_year = min(year),
    max_year = max(year),
    n_families = n_distinct(family)
  )
```

## Support diagnostics

```{r support-plots}
samples_by_year_method <- sample_meta_main |>
  count(year, collection_method, name = "n_samples")

samples_by_colony_method <- sample_meta_main |>
  count(colony, collection_method, name = "n_samples")

sample_map <- sample_meta_main |>
  ggplot(aes(lon, lat, colour = collection_method)) +
  geom_point(alpha = 0.8, size = 2) +
  coord_sf() +
  labs(
    title = "Retained samples in the main analysis period",
    subtitle = "Colonies through Onsala only; methods restricted to Faeces and Hunt",
    x = "Longitude",
    y = "Latitude",
    colour = "Method"
  )

plot_year_method <- samples_by_year_method |>
  ggplot(aes(year, n_samples, fill = collection_method)) +
  geom_col(position = "stack") +
  labs(
    title = "Samples by year and collection method",
    x = "Year",
    y = "Samples",
    fill = "Method"
  )

plot_colony_method <- samples_by_colony_method |>
  ggplot(aes(colony, n_samples, fill = collection_method)) +
  geom_col(position = "stack") +
  labs(
    title = "Samples by colony and collection method",
    subtitle = "Colony is retained for support diagnostics only",
    x = NULL,
    y = "Samples",
    fill = "Method"
  )

plot_family_support <- family_support_grouped |>
  mutate(family = fct_reorder(as.character(family), total_biomass_kg)) |>
  ggplot(aes(total_biomass_kg, family)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Grouped prey families retained for modelling",
    subtitle = "Fixed dominant-family set; all remaining families grouped as Other",
    x = "Total biomass (kg)",
    y = NULL
  )

plot_zero_rate <- family_support_grouped |>
  mutate(family = fct_reorder(as.character(family), zero_rate)) |>
  ggplot(aes(zero_rate, family)) +
  geom_col(fill = "tomato") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Zero proportion by prey family",
    x = "Zero proportion",
    y = NULL
  )

sample_map
plot_year_method
plot_colony_method
plot_family_support
plot_zero_rate

#ggsave(file.path(figures_dir, "sample-map-main-period.png"), sample_map, width = 8, height = 6, dpi = 300)
#ggsave(file.path(figures_dir, "samples-by-year-method.png"), plot_year_method, width = 9, height = 5, dpi = 300)
#ggsave(file.path(figures_dir, "samples-by-colony-method.png"), plot_colony_method, width = 8, height = 5, dpi = 300)
#ggsave(file.path(figures_dir, "family-support-grouped.png"), plot_family_support, width = 8, height = 5, dpi = 300)
#ggsave(file.path(figures_dir, "family-zero-rate.png"), plot_zero_rate, width = 8, height = 5, dpi = 300)
```

# <--- Max continued from here (above is from script 03)

```{r}
# Gridded map prediction
d <- diet_model

x <- d$X
y <- d$Y
z <- chull(x, y)

coords <- cbind(x[z], y[z])

coords <- rbind(coords, coords[1, ])

plot(coords[, 1] ~ coords[, 2]) # plot data

sp_poly <- sp::SpatialPolygons(
  list(sp::Polygons(list(sp::Polygon(coords)), ID = 1))
)

sp_poly_df <- sp::SpatialPolygonsDataFrame(sp_poly,
  data = data.frame(ID = 1)
)
cell_width <- 0.5

pred_grid <- expand.grid(
  X = seq(min(d$X), max(d$X), cell_width),
  Y = seq(min(d$Y), max(d$Y), cell_width),
  year = c(min(d$year):max(d$year))
)

sp::coordinates(pred_grid) <- c("X", "Y")

inside <- !is.na(sp::over(pred_grid, as(sp_poly_df, "SpatialPolygons")))

pred_grid <- pred_grid[inside, ]

pred_grid <- as.data.frame(pred_grid) |> 
  mutate(collection_method = "Hunt",
         year_ct = year - year_center,
         year_f = as.factor(year))

pred_grid <- replicate_df(pred_grid,
    time_name = "family",
    time_values = unique(d$family)
    )
```

## Map plot

```{r}
# Add coastline
land <- ne_countries(scale = "large", returnclass = "sf") |>
  st_transform(32632)
```

## Mesh

```{r mesh}
mesh_cutoff_km <- 0.5

mesh <- make_mesh(diet_model, xy_cols = c("X", "Y"), type = "cutoff", cutoff = mesh_cutoff_km)

plot(mesh)
```

## Candidate models

```{r fit-models}
#| cache: true
# Spatial no time
m1 <- sdmTMB(
  formula = biomass_kg ~ 0 + family + collection_method,
  data = diet_model,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(),
  mesh = mesh,
  spatial = "on"
)

# smooth year by family
m2 <- sdmTMB(
  formula = biomass_kg ~ 0 + family + collection_method + s(year_ct, by = family),
  data = diet_model,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(),
  mesh = mesh,
  spatial = "on"
)

# SVC family + year smooth
m3 <- sdmTMB(
  formula = biomass_kg ~ 0 + family + collection_method + s(year_ct, by = family),
  data = diet_model,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  spatial_varying = ~0 + family
)

# SVC family + year smooth + iid
m4 <- sdmTMB(
  formula = biomass_kg ~ 0 + family + collection_method + s(year_ct, by = family),
  data = diet_model,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  time = "year",
  spatiotemporal = "iid",
  spatial_varying = ~0 + family
)

# factor year by family, collection by family, no st
m5 <- sdmTMB(
  formula = biomass_kg ~ 0 + family + collection_method*family + year_f*family,
  data = diet_model,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  time = "year",
  spatiotemporal = "off",
  spatial_varying = ~0 + family
)

# factor year by family, collection by family, iid st
m6 <- sdmTMB(
  formula = biomass_kg ~ 0 + family + collection_method*family + year_f*family,
  data = diet_model,
  #family = delta_gamma(type = "poisson-link"),
  family = tweedie(),
  mesh = mesh,
  spatial = "off",
  time = "year",
  spatiotemporal = "iid",
  spatial_varying = ~0 + family
)

# SVC family, random walk time by family
AIC(m1, m2, m3, m4, m5, m6) |> arrange(AIC)
```

```{r}
m6
sanity(m6)

# Predict over time
ndt <- tibble(
  expand_grid(
    #year = modelr::seq_range(diet_model$year, n = 50), # for smooth!
    year = unique(diet_model$year),
    family = unique(diet_model$family)))  |> 
  mutate(collection_method = "Hunt",
         #year_ct = year - year_center,
         year_f = as.factor(year),
         year = round(year))
  
p <- predict(m6, newdata = ndt, re_form = NA, se_fit = TRUE)

tot <- p |>
  mutate(sum = sum(est), .by = year) |> 
  filter(family == "Other") |> 
  mutate(family = "Total",
         est_se = NA)


p |> 
  bind_rows(tot) |> 
  ggplot(aes(year, exp(est), color = family, fill = family)) + 
  geom_ribbon(aes(ymin = exp(est - est_se*1.96),
                    ymax = exp(est + est_se*1.96)),
              alpha = 0.3, color = NA) +
  scale_x_continuous(breaks = seq(2012, 2025, by = 3)) +
  scale_color_manual(values = family_palette) +
  scale_fill_manual(values = family_palette) +
  theme(legend.position = "right") +
  guides(fill = "none") +
  labs(y = "Kg/sample", color = "Family") +
  facet_wrap(~family, scales = "free") +
  geom_line()

# Plot proportion over time
p |> 
  mutate(sum = sum(exp(est)), 
         prop = exp(est)/sum,
         .by = c(year)) |> 
  ggplot(aes(year, prop, color = family, fill = family)) + 
  scale_x_continuous(breaks = seq(2012, 2025, by = 3)) +
  scale_color_manual(values = family_palette) +
  scale_fill_manual(values = family_palette) +
  coord_cartesian(expand = 0) +
  geom_area()
```

```{r}
# Predict!
pg <- predict(m6, newdata = pred_grid)

# Plot SVCs
pg |> 
  filter(year == 2013) |> 
  pivot_longer(c("zeta_s_familyGadidae", "zeta_s_familyClupeidae",
                 "zeta_s_familyPleuronectidae",
                 "zeta_s_familyGobiidae", "zeta_s_familyOther"),
               names_to = "var") |> 
  mutate(var = str_remove(var, "zeta_s_family")) |> 
  ggplot(aes(X*1000, Y*1000, fill = value)) +
  geom_raster() +
  facet_wrap(~var, ncol = 5) +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
    coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2()

# Plot spatial predictions in a given year
pg |> 
  filter(year == 2013) |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~family, ncol = 5)

# Now do by year for cod
pg |> 
  filter(family == "Gadidae") |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  guides(fill = guide_colorbar(position = "inside")) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position.inside = c(0.95, 0.2)) +
  facet_wrap(~year, ncol = 7)

# ... and clupeids
pg |> 
  filter(family == "Clupeidae") |> 
  ggplot(aes(X*1000, Y*1000, fill = exp(est))) +
  geom_raster() +
  geom_sf(data = land, fill = "grey90", linewidth = 0.2, inherit.aes = FALSE) +
  viridis::scale_fill_viridis(trans = "sqrt") +
  coord_sf(
    crs = 32632,
    xlim = range(pg$X*1000),
    ylim = range(pg$Y*1000),
    expand = FALSE
  ) +
  labs(x = "Longitude", y = "Latitude") +
  guides(fill = guide_colorbar(position = "inside")) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position.inside = c(0.95, 0.2)) +
  facet_wrap(~year, ncol = 7)
```

Plot effect of collection method:

```{r}
nd_col <- tibble(
  expand_grid(
    collection_method = unique(diet_model$collection_method),
    family = unique(diet_model$family)
  )) |> 
  mutate(year = 2013,
         year_f = as.factor(year))

pcol <- predict(m6, newdata = nd_col, se_fit = TRUE, re_form = NA)

ggplot(pcol, aes(exp(est), family, color = collection_method)) + 
  geom_point(position = position_dodge(width = 0.2)) + 
  geom_errorbar(aes(xmin = exp(est - 1.96*est_se),
                    xmax = exp(est + 1.96*est_se)),
                width = 0,
                position = position_dodge(width = 0.2))
```






 