---
title: "Seal Exploration"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    page-layout: full
    embed-resources: true
editor: source
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
---

## Purpose

This notebook explores harbour seal body length data for the same filtered sample subset used in `03b-diet-models-family.qmd`.

The workflow is:

- retain colonies through Onsala only,
- retain Faeces and Hunt samples only,
- retain samples with positive prey biomass,
- retain 2013-2025 as the main analysis period,
- summarise seal length availability,
- plot the seal length frequency distribution,
- check whether seal size varies over time or space.

```{r setup}
library(dplyr)
library(readr)
library(ggplot2)
library(janitor)
library(ggsidekick)

theme_set(theme_sleek())

home <- here::here()
biomass_file <- file.path(home, "Data", "HS_SKAG_estimated_abundance_biomass_130426.csv")

colony_keep <- c("Koster", "Väderöarna", "Lysekil", "Marstrand", "Onsala")
method_keep <- c("Faeces", "Hunt")
analysis_year_min <- 2013L

normalize_colony_name <- function(x) {
  x <- trimws(as.character(x))
  case_when(
    grepl("der.*arna", x, ignore.case = TRUE) ~ "Väderöarna",
    grepl("^Hallands.*der", x, ignore.case = TRUE) ~ "Hallands Väderö",
    TRUE ~ x
  )
}
```

## Read and filter sample metadata

```{r read-data}
seal_raw <- read_csv(biomass_file, show_col_types = FALSE) |>
  clean_names()

sample_meta_all <- seal_raw |>
  distinct(sample_id, .keep_all = TRUE) |>
  transmute(
    sample_id,
    year = suppressWarnings(as.integer(year)),
    month = suppressWarnings(as.integer(month)),
    quarter = coalesce(suppressWarnings(as.integer(quarter)), ((month - 1L) %/% 3L) + 1L),
    lat = suppressWarnings(as.numeric(latitude)),
    lon = suppressWarnings(as.numeric(longitude)),
    colony = normalize_colony_name(colony),
    collection_method = if_else(
      is.na(collection_method) | collection_method == "",
      "Unknown",
      as.character(collection_method)
    ),
    body_length_cm = suppressWarnings(as.numeric(body_length_cm))
  )

sample_totals_all <- seal_raw |>
  filter(
    !is.na(sample_id),
    !is.na(biomass_not_corrected),
    !is.na(prey_species_lat),
    prey_species_lat != "",
    tolower(prey_species_lat) != "non prey item"
  ) |>
  transmute(
    sample_id,
    biomass_kg = biomass_not_corrected / 1000
  ) |>
  summarise(sum_biomass_kg = sum(biomass_kg, na.rm = TRUE), .by = sample_id)

sample_meta_main <- sample_meta_all |>
  left_join(sample_totals_all, by = "sample_id") |>
  mutate(sum_biomass_kg = coalesce(sum_biomass_kg, 0)) |>
  filter(
    !is.na(year),
    !is.na(lat),
    !is.na(lon),
    collection_method %in% method_keep,
    colony %in% colony_keep,
    sum_biomass_kg > 0,
    year >= analysis_year_min
  )
```

## Seal length availability

```{r length-summary}
sample_meta_main |>
  summarise(
    total_samples = n(),
    samples_with_body_length = sum(!is.na(body_length_cm)),
    prop_with_body_length = samples_with_body_length / total_samples,
    mean_body_length_cm = mean(body_length_cm, na.rm = TRUE),
    sd_body_length_cm = sd(body_length_cm, na.rm = TRUE),
    min_body_length_cm = min(body_length_cm, na.rm = TRUE),
    median_body_length_cm = median(body_length_cm, na.rm = TRUE),
    max_body_length_cm = max(body_length_cm, na.rm = TRUE)
  )

sample_meta_main |>
  count(collection_method, has_body_length = !is.na(body_length_cm))
```

## Seal length frequency distribution

```{r length-frequency}
sample_meta_main |>
  filter(!is.na(body_length_cm)) |>
  ggplot(aes(x = body_length_cm)) +
  geom_histogram(binwidth = 5, color = "white", fill = "grey40") +
  labs(
    x = "Seal length (cm)",
    y = "Number of samples",
    title = "Seal length frequency distribution",
    subtitle = "03b analysis subset"
  )
```

## Trend over time

```{r length-over-time}
seal_length_data <- sample_meta_main |>
  filter(!is.na(body_length_cm))

seal_length_data |>
  summarise(
    cor_year_length = cor(year, body_length_cm),
    n = n()
  )

seal_length_data |>
  summarise(
    mean_length_cm = mean(body_length_cm),
    .by = year
  ) |>
  ggplot(aes(year, mean_length_cm)) +
  geom_point(size = 2) +
  geom_line() +
  labs(
    x = "Year",
    y = "Mean seal length (cm)",
    title = "Mean seal length by year",
    subtitle = "03b analysis subset"
  )

seal_length_data |>
  ggplot(aes(year, body_length_cm)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    x = "Year",
    y = "Seal length (cm)",
    title = "Seal length over time",
    subtitle = "Linear trend shown for reference"
  )

summary(lm(body_length_cm ~ year, data = seal_length_data))$coefficients
```

## Pattern in space

```{r length-in-space}
seal_length_data |>
  summarise(
    n = n(),
    mean_length_cm = mean(body_length_cm),
    sd_length_cm = sd(body_length_cm),
    .by = colony
  ) |>
  arrange(desc(n))

seal_length_data |>
  ggplot(aes(colony, body_length_cm)) +
  geom_boxplot() +
  labs(
    x = NULL,
    y = "Seal length (cm)",
    title = "Seal length by colony",
    subtitle = "03b analysis subset"
  )

seal_length_data |>
  ggplot(aes(lon, lat, color = body_length_cm)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "C") +
  coord_equal() +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Length (cm)",
    title = "Spatial distribution of seal length",
    subtitle = "03b analysis subset"
  )

summary(lm(body_length_cm ~ lon + lat, data = seal_length_data))$coefficients
```
