---
title: "Survey Fish Availability: Distribution Maps"
author: "Monica Mion & Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

## Purpose

The modelling choice is deliberately simple:

- restrict to Q3 only,
- do not include year effects,
- do not include quarter effects,
- compare a non-spatial model to a spatial model,
- map the best-supported Q3 spatial surface.


## Load libraries

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sdmTMB)
library(viridis)
library(sf)
library(sp)
library(terra)

theme_set(theme_light())

home <- here::here()
```

## Read family data

```{r}
dat <- read_csv(paste0(home, "/Data/survey/survey_family_model_data.csv"), show_col_types = FALSE) |>
  mutate(
    survey = factor(survey),
    quarter = factor(quarter),
    family = factor(family)
  )
```

## Restrict to quarter 3

```{r}
#| echo: false
dat_q3 <- dat |>
  filter(quarter == 3)

dat_q3 |>
  distinct(haul_id, survey, year) |>
  count(survey)
```

This restriction removes the main survey-quarter confounding problem and keeps the comparison focused on the period where both surveys sample the system.

## Family support in quarter 3

```{r}
#| echo: false
family_support <- dat_q3 |>
  group_by(family) |>
  summarise(
    total = sum(catch_kg_km2),
    n_hauls = n_distinct(haul_id),
    zero_rate = mean(catch_kg_km2 == 0),
    surveys = paste(sort(unique(survey)), collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(desc(total))

family_support
```

## Families selected for mapping

We focus on:

- Gadidae joint IBTS + CTS Q3 map
- Pleuronectidae joint IBTS + CTS Q3 map
- Clupeidae IBTS-only Q3 map

## Build a prediction grid

```{r}
#| echo: false
haul_pts <- dat_q3 |>
  distinct(haul_id, survey, X, Y, lat, lon)

x <- haul_pts$X
y <- haul_pts$Y
z <- chull(x, y)
coords <- cbind(x[z], y[z])
coords <- rbind(coords, coords[1, ])

sp_poly <- sp::SpatialPolygons(
  list(sp::Polygons(list(sp::Polygon(coords)), ID = 1))
)
sp_poly_df <- sp::SpatialPolygonsDataFrame(sp_poly, data = data.frame(ID = 1))

cell_width <- 2

pred_grid <- expand.grid(
  X = seq(min(haul_pts$X), max(haul_pts$X), cell_width),
  Y = seq(min(haul_pts$Y), max(haul_pts$Y), cell_width)
)

sp::coordinates(pred_grid) <- c("X", "Y")
inside <- !is.na(sp::over(pred_grid, as(sp_poly_df, "SpatialPolygons")))
pred_grid <- as.data.frame(pred_grid[inside, ])

xy <- as.matrix(pred_grid |> select(X, Y) |> mutate(X = X * 1000, Y = Y * 1000))
v <- terra::vect(xy, crs = "+proj=utm +zone=33 +datum=WGS84 +units=m")
ll <- terra::project(v, "+proj=longlat +datum=WGS84")
pred_grid$lon <- terra::geom(ll)[, "x"]
pred_grid$lat <- terra::geom(ll)[, "y"]
pred_grid$quarter <- factor(3, levels = levels(dat_q3$quarter))
pred_grid$survey <- factor("IBTS", levels = levels(dat_q3$survey))
```

## Prediction grid check

```{r}
#| echo: false
ggplot(pred_grid, aes(X, Y)) +
  geom_point(size = 0.1) +
  coord_equal() +
  labs(
    title = "Prediction grid",
    subtitle = "2 x 2 km grid inside the convex hull of Q3 hauls",
    x = "UTM Easting (km)",
    y = "UTM Northing (km)"
  )
```

## Helper functions

```{r}
fit_q3_models <- function(d, use_survey = TRUE) {
  mesh <- make_mesh(d, xy_cols = c("X", "Y"), cutoff = 5)

  if (use_survey) {
    formula <- catch_kg_km2 ~ survey
  } else {
    formula <- catch_kg_km2 ~ 1
  }

  m1 <- sdmTMB(
    formula,
    data = d,
    mesh = mesh,
    family = tweedie(link = "log"),
    spatial = "off",
    spatiotemporal = "off",
    anisotropy = FALSE
  )

  m2 <- sdmTMB(
    formula,
    data = d,
    mesh = mesh,
    family = tweedie(link = "log"),
    spatial = "on",
    spatiotemporal = "off",
    anisotropy = FALSE
  )

  list(mesh = mesh, m1 = m1, m2 = m2)
}

get_aic_table <- function(model_list) {
  AIC(model_list$m1, model_list$m2) |>
    as_tibble(rownames = "model") |>
    arrange(AIC)
}

plot_family_surface <- function(pred, title, subtitle) {
  ggplot(pred, aes(X, Y, fill = est)) +
    geom_raster() +
    coord_equal() +
    scale_fill_viridis_c() +
    labs(
      title = title,
      subtitle = subtitle,
      x = "UTM Easting (km)",
      y = "UTM Northing (km)",
      fill = "kg/km2"
    )
}
```

## Gadidae: Q3 distribution map

```{r}
#| echo: false
gadidae_dat <- dat_q3 |>
  filter(family == "Gadidae")

gadidae_models <- fit_q3_models(gadidae_dat, use_survey = TRUE)
get_aic_table(gadidae_models)
```

```{r}
#| echo: false
sanity(gadidae_models$m2)
```

```{r}
#| echo: false
plot(gadidae_models$mesh$mesh, asp = 1)
gad_pts <- gadidae_dat |> distinct(haul_id, X, Y)
points(gad_pts$X, gad_pts$Y, pch = 16, cex = 0.3, col = scales::alpha("black", 0.4))
```

```{r}
#| echo: false
pred_gadidae <- pred_grid |>
  mutate(survey = factor("IBTS", levels = levels(gadidae_dat$survey)))

gadidae_pred <- predict(gadidae_models$m2, newdata = pred_gadidae)

plot_family_surface(
  gadidae_pred,
  title = "Gadidae distribution in Q3",
  subtitle = "Spatial Tweedie model; IBTS reference surface"
)
```

## Pleuronectidae: Q3 distribution map

```{r}
#| echo: false
pleur_dat <- dat_q3 |>
  filter(family == "Pleuronectidae")

pleur_models <- fit_q3_models(pleur_dat, use_survey = TRUE)
get_aic_table(pleur_models)
```

```{r}
#| echo: false
sanity(pleur_models$m2)
```

```{r}
#| echo: false
pred_pleur <- pred_grid |>
  mutate(survey = factor("IBTS", levels = levels(pleur_dat$survey)))

pleur_pred <- predict(pleur_models$m2, newdata = pred_pleur)

plot_family_surface(
  pleur_pred,
  title = "Pleuronectidae distribution in Q3",
  subtitle = "Spatial Tweedie model; IBTS reference surface"
)
```

## Clupeidae: Q3 IBTS-only map

```{r}
#| echo: false
clup_dat <- dat_q3 |>
  filter(family == "Clupeidae", survey == "IBTS")

clup_models <- fit_q3_models(clup_dat, use_survey = FALSE)
get_aic_table(clup_models)
```

```{r}
#| echo: false
sanity(clup_models$m2)
```

```{r}
#| echo: false
clup_pred <- predict(clup_models$m2, newdata = pred_grid)

plot_family_surface(
  clup_pred,
  title = "Clupeidae distribution in Q3 (IBTS only)",
  subtitle = "Spatial Tweedie model"
)
```

## Next step

1. add ICES/land background layers to these maps,
2. compare these Q3 family maps to harbour seal diet composition,
3. build a species-level Q3 version for cod, whiting, haddock, plaice, and herring.
