---
title: "Prepare raw diet data"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

## Load packages and set working directory

```{r}
#| message: false
# Install libraries if not already installed using install.packages("package_name")
library(dplyr)

home <- here::here()
```

## Load data

```{r READ RAW DATA}
data <- read.csv(paste0(home, "/Data/HS_SKAG_raw data_stomachs.csv"), fileEncoding = "UTF-8")
## Replace empty strings with NA
data <- data |>
  mutate(across(where(is.character), ~ na_if(., "")))

## Standardize prey remain labels
data <- data |>
  mutate(  
    Prey.remain = tolower(trimws(Prey.remain))
  )
```

## Load regression tables
```{r}
reg_FL <- read.csv(paste0(home, "/Data/Regression-otolith/Regressions_otolith_FL.csv"))
reg_FW <- read.csv(paste0(home,"/Data/Regression-otolith/Regressions_otolith_FW.csv"))

## Harmonize taxon column
reg_FL <- reg_FL |> rename(Prey.species..lat. = Form_Taxon)
reg_FW <- reg_FW |> rename(Prey.species..lat. = Form_Taxon)
```

## Prepare length regressions
```{r}
reg_OW_FL <- reg_FL |>
  filter(Goal_unit == "Lenght (FL)", Count_variabel == "OW") |>
  select(Prey.species..lat., a_OW_FL = a, b_OW_FL = b)

reg_OL_FL <- reg_FL |>
  filter(Goal_unit == "Lenght (FL)", Count_variabel == "OL") |>
  select(Prey.species..lat., a_OL_FL = a, b_OL_FL = b)
```

## Join length regressions
```{r}
data <- data |>
  left_join(reg_OW_FL, by = "Prey.species..lat.") |>
  left_join(reg_OL_FL, by = "Prey.species..lat.")
## Ensure numeric otolith values
data <- data |>
  mutate(
    Otolith.width..mm.   = as.numeric(Otolith.width..mm.),
    Otolith.length...mm. = as.numeric(Otolith.length...mm.)
  )
```

## Calculate fish length (FL_not_corrected)

```{r}
data <- data |>
  mutate(
    FL_not_corrected = case_when(
      !is.na(Otolith.width..mm.) &
        !is.na(a_OW_FL) & !is.na(b_OW_FL) ~
        a_OW_FL + b_OW_FL * Otolith.width..mm.,
      
      is.na(Otolith.width..mm.) &
        !is.na(Otolith.length...mm.) &
        !is.na(a_OL_FL) & !is.na(b_OL_FL) ~
        a_OL_FL + b_OL_FL * Otolith.length...mm.,
      
      TRUE ~ NA_real_
    )
  )
```

## Prepare weight regressions

```{r}
reg_OW_FW <- reg_FW |>
  filter(Goal_unit == "Weight (FW)", Count_variabel == "OW") |>
  select(Prey.species..lat., a_OW_FW = a, b_OW_FW = b, Regression_OW = Regression)

reg_OL_FW <- reg_FW |>
  filter(Goal_unit == "Weight (FW)", Count_variabel == "OL") |>
  select(Prey.species..lat., a_OL_FW = a, b_OL_FW = b, Regression_OL = Regression)

reg_FL_FW <- reg_FW |>
  filter(Goal_unit == "Weight (FW)", Count_variabel == "FL") |>
  select(Prey.species..lat., a_FL_FW = a, b_FL_FW = b, Regression_FL = Regression)
```

## Join weight regressions

```{r}
data <- data |>
  left_join(reg_OW_FW, by = "Prey.species..lat.") |>
  left_join(reg_OL_FW, by = "Prey.species..lat.") |>
  left_join(reg_FL_FW, by = "Prey.species..lat.")
```

## DEFINE AVAILABLE MEASURED FL

```{r}
data <- data |>
  mutate(
    all_FL_available_ok = case_when(
      Prey.remain %in% c("whole fish", "whole body", "whole individual") &
        !is.na(Prey.length.measured..cm.) ~
        as.numeric(Prey.length.measured..cm.),
      TRUE ~ NA_real_
    )
  )
```

## CALCULATE FISH WEIGHT (FW_not_corrected)

```{r}
data <- data |>
  mutate(
    FW_not_corrected = case_when(
      
      ## Stickleback fixed mass
      Family == "Gasterosteidae" ~ 1.22,
      
      ## Otolith width regressions
      !is.na(Otolith.width..mm.) &
        Regression_OW == "Y=(aX)^b" ~
        (a_OW_FW * Otolith.width..mm.)^b_OW_FW,
      
      !is.na(Otolith.width..mm.) &
        Regression_OW == "Y=aX^b" ~
        a_OW_FW * Otolith.width..mm.^b_OW_FW,
      
      !is.na(Otolith.width..mm.) &
        Regression_OW == "Y=a+bX" ~
        a_OW_FW + b_OW_FW * Otolith.width..mm.,
      
      ## Otolith length regressions
      is.na(Otolith.width..mm.) &
        !is.na(Otolith.length...mm.) &
        Regression_OL == "Y=(aX)^b" ~
        (a_OL_FW * Otolith.length...mm.)^b_OL_FW,
      
      is.na(Otolith.width..mm.) &
        !is.na(Otolith.length...mm.) &
        Regression_OL == "Y=aX^b" ~
        a_OL_FW * Otolith.length...mm.^b_OL_FW,
      
      is.na(Otolith.width..mm.) &
        !is.na(Otolith.length...mm.) &
        Regression_OL == "Y=a+bX" ~
        a_OL_FW + b_OL_FW * Otolith.length...mm.,
      
      ## Measured FL regressions
      !is.na(all_FL_available_ok) &
        Regression_FL == "W(g)=(a*10^-6* L(cm)^b) *1000" ~
        (a_FL_FW * 1e-6 * (all_FL_available_ok^b_FL_FW)) * 1000,
      
      !is.na(all_FL_available_ok) &
        Regression_FL %in% c("Y=aX^b", "Y=(aX)^b") ~
        a_FL_FW * (all_FL_available_ok^b_FL_FW),
      
      TRUE ~ NA_real_
    )
  )
```

## Calculate abundance

```{r}
data <- data |>
  mutate(
    n = as.numeric(n),
    abundance = case_when(
      Prey.remain %in% c("otolith", "broken otolith") ~ n / 2,
      Prey.remain == "spine" & Family == "Gasterosteidae" ~ n / 5,
      Prey.remain %in% c(
        "whole fish", "whole individual", "fragment", "bones",
        "green bones", "plastic", "stone", "scales",
        "dermal plaques", "teeth", "eggs"
      ) ~ n,
      TRUE ~ NA_real_
    )
  )
```

## Calculate biomass

```{r}
data <- data |>
  mutate(
    biomass_not_corrected = FW_not_corrected * abundance
  )

write.csv(
  data,paste0(home, "/Data/HS_SKAG_estimated_abundance_biomass.csv"),
  row.names = FALSE
)
```

## Quick checks

```{r}
summary(data$FL_not_corrected)
summary(data$FW_not_corrected)
summary(data$biomass_not_corrected)
```

## Check for missing regressions

```{r}
## --- Unique prey taxa in stomach data 

prey_data <- data |>
  distinct(Prey.species..lat.)

## --- Available regressions: LENGTH 

reg_length_available <- reg_FL |>
  filter(Goal_unit == "Lenght (FL)") |>
  distinct(Prey.species..lat.)

## --- Available regressions: WEIGHT -----------------------

reg_weight_available <- reg_FW |>
  filter(Goal_unit == "Weight (FW)") |>
  distinct(Prey.species..lat.)

## --- Combine checks --------------------------------------

regression_check <- prey_data |>
  left_join(
    reg_length_available |> mutate(has_FL_regression = TRUE),
    by = "Prey.species..lat."
  ) |>
  left_join(
    reg_weight_available |> mutate(has_FW_regression = TRUE),
    by = "Prey.species..lat."
  ) |>
  mutate(
    has_FL_regression = ifelse(is.na(has_FL_regression), FALSE, has_FL_regression),
    has_FW_regression = ifelse(is.na(has_FW_regression), FALSE, has_FW_regression)
  )

## --- Summary flags 

regression_check <- regression_check |>
  mutate(
    missing_FL = !has_FL_regression,
    missing_FW = !has_FW_regression,
    missing_both = missing_FL & missing_FW
  )

## --- RESULTS 

## Species missing ANY regression
missing_any <- regression_check |>
  filter(missing_FL | missing_FW)

## Species missing BOTH length and weight regressions
missing_both <- regression_check |>
  filter(missing_both)

## Species with full coverage
complete_coverage <- regression_check |>
  filter(!missing_FL & !missing_FW)

############################################
# PRINT RESULTS
############################################

cat("\n--- Species missing ANY regression ---\n")
print(missing_any)

cat("\n--- Species missing BOTH FL and FW regressions ---\n")
print(missing_both)

cat("\n--- Species with complete regression coverage ---\n")
print(complete_coverage)

```

