---
title: "Survey data preparation"
author: "Max Lindmark & Monica Mion"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    toc: true
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 80%
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

## Load libraries

```{r}
#| message: false
library(tidyverse)
library(sf)

  home <- here::here()

species_family_path <- paste0(home, "/Data/survey/species_family.csv")
ibts_path <- paste0(home, "/Data/survey/Species Per Fishing Operation_IBTS_1999-2026.csv")
ct_path <- paste0(home, "/Data/survey/coast_trawl_2013_2024.csv")
out_path <- paste0(home, "/Data/survey/survey_family_model_data.csv")
species_out_path <- paste0(home, "/Data/survey/survey_species_model_data.csv")
```

# PART 1: The Two Surveys

**Time span we use:** 2013–2024 (the overlap with the coastal trawl).

**Known issue:** IBTS: The Swept area (km2) column is sometimes missing. When it is, we estimate it using the distance towed × door spread of the GOV net (standard GOV geometry: distance in nautical miles × 1.852 × 0.0862 km²).

CTS: Coordinates are in degrees-minutes format (e.g. "5837.859" means 58°37.859'). We must convert these to decimal degrees (e.g. 58.631°) before using them.

# PART 3: The Species → Family lookup Table

## STEP 1: Load the lookup table
Taxonomic list to add column family from species names. This is the same file used in the seal diet analysis, so we can be sure the family names match.

```{r}
species_family <- read_csv(species_family_path)
```

## STEP 2: Load IBTS data

```{r}
ibts_raw <- read_csv(ibts_path)
```

------------------------------------------------------------------------

## STEP 3: Process IBTS

### 3a. Filter to what we actually need

```{r}
ibts_filt <- ibts_raw |>
  filter(
    `Station validity` == "Valid",   # Only valid hauls
    Subdiv == 20,                    # Only Skagerrak (not North Sea)
    Year >= 2013, Year <= 2024       # Overlap years with CTS
  )
```

This removes: Invalid or incomplete hauls, Hauls outside the Skagerrak, Years outside 2013–2024

### 3b. Calculate swept area

```{r}
ibts_filt <- ibts_filt |>
  mutate(
    year = Year,
    quarter = Quarter,
    lat = `Lat. stop dec`,
    lon = `Long. stop dec`,
    swept_area = coalesce(
      `Swept area (km2)`,           # Use the existing column if available
      `Dist (Nm)` * 1.852 * 0.0862  # Else estimate: distance × conversion factor
    )
  )
```

coalesce() means: *"use the first value that is not NA; if it IS NA, use the second value."*


### 3c. Keep only hauls with valid swept area

```{r}
haul_meta <- ibts_filt |>
  filter(!is.na(swept_area), swept_area > 0, !is.na(lat), !is.na(lon)) |>
  distinct(year, quarter, Vessel, `Trip number`, Haul, lat, lon)
```

### 3d. Normalize species names

```{r}
ibts_filt <- ibts_filt |>
  mutate(specname_norm = tolower(gsub(" ", "_", `Specname sci.`)))
```
This step makes species taxonomic file and survey match.

### 3e. Map species → family

```{r}
ibts_filt <- ibts_filt |>
  left_join(species_family, by = c("specname_norm" = "specname_sci"))
```

For each row in IBTS, we look up what family the species belongs to. 

### 3f. Calculate catch density

This means: *"How many kilograms of fish were found per square kilometre of seabed?"*

```{r}
ibts_filt <- ibts_filt |>
  mutate(catch_kg_km2 = `Catch weight (kg)` / swept_area)
```

This means: *"How many kilograms of fish were found per square kilometre of seabed?"*

This is important because a 30-minute haul covers more area than a 10-minute haul. Raw catch weight would be misleading — longer hauls would always show more fish just because they sampled more seabed. Density normalises for this.

### 3g. Aggregate to haul × family level

```{r}
ibts_haul <- ibts_filt |>
  filter(!is.na(family)) |>
  group_by(Year, Quarter, `Trip number`, Haul, Vessel,
           `Lat. stop dec`, `Long. stop dec`, family) |>
  summarise(catch_kg_km2 = sum(catch_kg_km2, na.rm = TRUE),
            .groups = "drop")
```

**What this does:** A single haul can catch multiple species of the same family. We sum them all up so we have **one row per haul per family**.

For example, a haul might catch: - 2 kg of cod (*Gadus morhua*, family = Gadidae) - 1.5 kg of haddock (*Melanogrammus aeglefinus*, family = Gadidae) - 0.5 kg of whiting (*Merlangius merlangus*, family = Gadidae)

After aggregation: - **Gadidae: 4 kg/km²** (sum of all three) - Clupeidae: 0 (nothing from that family) - etc.

### 3h. Create a unique haul ID

```{r}
ibts_haul <- ibts_haul |>
  mutate(
    haul_id = paste(Year, Quarter, Vessel, `Trip number`, Haul,
                    round(`Lat. stop dec`, 4), round(`Long. stop dec`, 4), sep = "_")
  )
```

We create a text label that uniquely identifies each haul: e.g. `"2019_3_SEPI_15_35_58.1234_11.5678"`. This is crucial for: - Joining IBTS and CTS data together later - Zero-filling (see Step 8) - Keeping track of hauls in the model

------------------------------------------------------------------------

## STEP 4 & 5: Load and Process Coast Trawl (CTS)

The CTS processing follows the same logic but with some differences:

### 4a. Convert deg-min coordinates to decimal degrees

```{r}
convert_dm <- function(x) {
  x <- as.character(x)
  deg  <- as.numeric(substr(x, 1, 2))
  mins <- as.numeric(substr(x, 3, nchar(x)))
  deg + mins / 60
}
```

The CTS file has coordinates in a weird format: `"5837.859"` means 58°37.859'.

Math: `58 + 37.859/60 = 58.631°`

We apply this function to the `Lat` and `Long` columns.

### 4b. Load and filter valid rows

```{r}
ct_raw <- read_csv(ct_path)

ct_filt <- ct_raw |>
  filter(Validity == "V", Year >= 2013, Year <= 2024) |>
  mutate(lat = convert_dm(Lat), lon = convert_dm(Long))
```

### 4c. Drop rows with missing kg_km2

```{r}
ct_filt <- ct_filt |>
  filter(!is.na(kg_km2), !is.na(lat), !is.na(lon))
```

Some species (like *Nephrops*) don't have length measurements → no weight → `kg_km2` is NA. These can't be used.

### 4d. Join species → family

```{r}
ct_filt <- ct_filt |>
  mutate(species = tolower(gsub(" ", "_", Scientific))) |>
  left_join(species_family, by = c("species" = "specname_sci")) |>
  filter(!is.na(family))
```

### 4e. Aggregate to haul × family

```{r}
ct_haul <- ct_filt |>

  group_by(Year, Quarter, Index, Haul, Station.name,
           lat, lon, Vattentyp, family) |>
  summarise(catch_kg_km2 = sum(kg_km2, na.rm = TRUE),
            .groups = "drop")
```

The CTS data has **one row per haul per species per length class**. A single haul at FLADEN might have 5 rows for the same species at different sizes. We sum them all up.

------------------------------------------------------------------------

## STEP 6: Combine Both Surveys

We combine species-level data from both surveys, then immediately aggregate to
family level for the spatial modelling pipeline.

**Species-level:** one row per haul × species (used for species analysis)
**Family-level:** one row per haul × family (used for family analysis and sdmTMB)

```{r}
surv <- bind_rows(
  ibts_filt |>
    filter(!is.na(family)) |>
    mutate(
      survey  = "IBTS",
      species = specname_norm,
      year    = Year,
      quarter = Quarter,
      haul_id = paste(Year, Quarter, Vessel, `Trip number`, Haul,
                      round(`Lat. stop dec`, 4), round(`Long. stop dec`, 4), sep = "_")
    ) |>
    select(year, quarter, haul_id, survey, lat, lon, species, family, catch_kg_km2),
  ct_filt |>
    filter(!is.na(family)) |>
    mutate(
      survey       = "CTS",
      catch_kg_km2 = kg_km2,
      year         = Year,
      quarter      = Quarter,
      haul_id = paste(Year, Quarter, Index, Haul,
                      round(lat, 4), round(lon, 4), sep = "_")
    ) |>
    select(year, quarter, haul_id, survey, lat, lon, species, family, catch_kg_km2)
)

surv_fam <- surv |>
  group_by(haul_id, year, quarter, survey, lat, lon, family) |>
  summarise(catch_kg_km2 = sum(catch_kg_km2, na.rm = TRUE), .groups = "drop")
```

We add a `Survey` column: "IBTS" for the international survey, "CTS" for the coastal survey.

------------------------------------------------------------------------

## STEP 7: Convert to UTM Coordinates

Apply to family-level data for the spatial modelling pipeline:

```{r}
surv_fam_sf <- st_as_sf(surv_fam, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
surv_fam_sf <- st_transform(surv_fam_sf, crs = 32633)
utm_coords  <- st_coordinates(surv_fam_sf)
surv_fam <- surv_fam |>
  mutate(X = round(utm_coords[, 1] / 1000, 3),
         Y = round(utm_coords[, 2] / 1000, 3))
```

**Why UTM?** sdmTMB needs coordinates in a projected coordinate system (not lat/lon). We use **UTM Zone 33** (EPSG:32633), which is the appropriate projection for the Skagerrak area.

-   `st_as_sf()`: turns the data into a spatial object with lat/lon coordinates
-   `st_transform()`: converts to UTM Zone 33
-   `st_coordinates()`: extracts the new X, Y values
-   `/ 1000`: converts metres → kilometres (sdmTMB prefers km)

**What do X and Y mean?** - **X** = Easting (horizontal distance from the UTM origin, in km) - **Y** = Northing (vertical distance from the UTM origin, in km)

Together, (X, Y) pinpoints each haul on a flat map — which is what sdmTMB needs to model spatial patterns.

## STEP 7b: Save Species-Level Dataset

`surv` now has species-level rows with individual species records. We add factor columns
and save to disk. Zero-filling (next step) only applies to the family-level analysis.

For species-level UTM coordinates, we transform separately:

```{r}
surv_sf <- st_as_sf(surv, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
surv_sf <- st_transform(surv_sf, crs = 32633)
utm_sp  <- st_coordinates(surv_sf)

surv_species <- surv |>
  mutate(X = round(utm_sp[, 1] / 1000, 3),
         Y = round(utm_sp[, 2] / 1000, 3)) |>
  mutate(
    year_f    = as.factor(year),
    year_ct   = year - min(year),
    quarter_f = as.factor(quarter),
    survey_f  = as.factor(survey),
    family_f  = as.factor(family),
    species_f = as.factor(species)
  ) |>
  select(haul_id, year, year_f, year_ct, quarter, quarter_f,
         survey, survey_f, lat, lon, X, Y,
         species, species_f, family, family_f, catch_kg_km2)

write_csv(surv_species, species_out_path)
```

------------------------------------------------------------------------

## STEP 8: Zero-Fill

First, define the full list of families observed in the data:

```{r}
all_families <- sort(unique(surv_fam$family))
```

Then expand so every haul has one row per family:

```{r}
surv_zf <- surv_fam |>
  group_by(haul_id) |>
  complete(family = all_families, fill = list(catch_kg_km2 = 0)) |>
  fill(year, quarter, survey, lat, lon, X, Y, .direction = "downup") |>
  ungroup()
```

sdmTMB needs every haul to have data for every family. A "0" means "this family was looked for but not found", it's different from "missing data."

This is exactly what the seal diet analysis already does (see `01-explore-data.R`), and it's standard practice in species distribution modelling.

------------------------------------------------------------------------

## STEP 9: Group Rare Families

We calculate what **percentage** of total fish biomass each family represents. Families contributing less than **3%** are grouped into "Other".

```{r}
total_bm <- sum(surv_zf$catch_kg_km2, na.rm = TRUE)
fam_share <- surv_zf |>
  group_by(family) |>
  summarise(fam_total = sum(catch_kg_km2, na.rm = TRUE), .groups = "drop") |>
  mutate(pct = 100 * fam_total / total_bm) |>
  arrange(desc(pct))

rare_families <- fam_share |>
  filter(pct < 3) |>
  pull(family)

surv_zf <- surv_zf |>
  mutate(family = if_else(family %in% rare_families, "Other", family)) |>
  group_by(haul_id, year, quarter, survey, lat, lon, X, Y, family) |>
  summarise(catch_kg_km2 = sum(catch_kg_km2, na.rm = TRUE), .groups = "drop")
```



## STEP 10: Final Data Types

```{r}
surv_final <- surv_zf |>
  mutate(
    year_f     = as.factor(year),
    year_ct    = year - min(year),
    quarter_f  = as.factor(quarter),
    survey_f   = as.factor(survey),
    family_f   = as.factor(family),
    catch_kg_km2 = replace_na(catch_kg_km2, 0)
  ) |>
  select(haul_id, year, year_f, year_ct, quarter, quarter_f,
         survey, survey_f, lat, lon, X, Y, family, family_f, catch_kg_km2)
```

We make sure all columns have the right data type for sdmTMB:

------------------------------------------------------------------------

## STEP 11: Save

The species-level dataset was saved earlier (step 3g). The family-level dataset
is saved here:

```{r}
write_csv(surv_final, out_path)
```

